Senescent cells inhibit muscle differentiation via the lipid- SASP 15d-PGJ2 mediated modification and control of HRas

Version 1

Abstract

Abstract

Senescent cells which are characterized by multiple features such as increased expression of SA β-gal and cell cycle inhibitors such as p21 or p16, accumulate with tissue damage and dysregulate tissue homeostasis. In the context of skeletal muscle, it is known that agents used for chemotherapy such as doxorubicin cause buildup of senescent cells, leading to the inhibition of tissue regeneration. Senescent cells influence neighboring cells via numerous secreted factors which form the senescence-associated secreted phenotype (SASP). Lipids are emerging as a key component of SASP that can control tissue homeostasis. Arachidonic acid-derived lipids have been shown to accumulate within senescent cells, specifically 15d-PGJ2, which is an electrophilic lipid produced by the non-enzymatic dehydration of the prostaglandin PGD2. In this study, we show that 15d-PGJ2 is also released by Doxorubicin-induced senescent cells as a SASP factor. Treatment of skeletal muscle myoblasts with the conditioned medium from these senescent cells inhibits myoblast fusion during differentiation. Inhibition of L-PTGDS, the enzyme that synthesizes PGD2, diminishes the release of 15d-PGJ2 by senescent cells and restores muscle differentiation. We further show that this lipid post-translationally modifies Cys184 of HRas in skeletal muscle cells, causing a reduction in the localization of HRas to the Golgi, increased HRas binding to RAF RBD, and activation of cellular MAPK-Erk signaling (but not the Akt signaling). Mutating C184 of HRas prevents the ability of 15d-PGJ2 to inhibit the differentiation of muscle cells and control the activity of HRas. This work shows that 15d-PGJ2 released from senescent cells could be targeted to restore muscle homeostasis after chemotherapy.

Introduction

Senescent cells are important drivers of aging and damage-associated loss of tissue homeostasis(Childs et al., 2015). Anti-cancer chemotherapy presents an important context where treatment with chemotherapeutics such as Doxorubicin (Doxo) causes widespread cellular senescence which inhibits tissue homeostasis and regeneration, including in skeletal muscles(Francis et al., 2022). It has been shown that Doxo causes systemic inflammation and leads to the emergence of senescent cells across tissues(Di Leonardo et al., 1994; Hu and Zhang, 2019; Robles and Adami, 1998). Senescent cells negatively affect tissue homeostasis and regeneration by releasing factors including proteins like growth factors, matrix metalloproteases, cytokines, and chemokines, and small molecules like fatty acid derivatives(Campisi, 2005; Coppé et al., 2010; Dilley et al., 2003; Krtolica et al., 2001; Parrinello et al., 2005; Shelton et al., 1999; Yang et al., 2006). The release of these factors from senescent cells is called the Senescence Associated Secretory Phenotype (SASP). It is expected that these SASP factors and their mechanisms of action will vary depending on cellular and tissue contexts. Identifying SASP factors and their underlying mechanistic targets will be critical for building an understanding of how senescent cells control tissue homeostasis(Coppé et al., 2010; Davalos et al., 2010). Lipids are a less explored family of SASP factors, and it is important to understand how they affect tissue regeneration(Hamsanathan and Gurkar, 2022). We have previously shown that senescent cells have increased intracellular levels of prostaglandin 15d-PGJ2(Wiley et al., 2021), a non-enzymatic dehydration product of prostaglandin PGD2(Shibata et al., 2002). In the context of skeletal muscle, PGD2 and 15d-PGJ2 have been shown to negatively regulate muscle differentiation via mechanisms that do not depend on a cognate receptor(Hunter et al., 2001; Veliça et al., 2010). Here we study the role of 15d-PGJ2 as a member of SASP and identify mechanisms by which it might negatively affect muscle regeneration. 15d-PGJ2 has been previously shown to covalently modify multiple proteins like MAPK1, MCM4, EIF4A-I, PKM1, GFAP etc. endothelial and neuronal cells(Marcone and Fitzgerald, 2013; Yamamoto et al., 2011). 15d-PGJ2 was shown to be covalently modifying HRas in NIH3T3, Cos1, and IMR90 cell lines(Luis Oliva et al., 2003; Wiley et al., 2021). We further studied HRas as an important target that might mediate the effects of 15d-PGJ2 on muscle differentiation via covalent modification. We investigated HRas as a possible effector of 15d-PGJ2 because (i) HRas belongs to the Ras superfamily of small molecule GTPases and is known regulator of key cellular processes(BOHR Orthopredic Hospital et al., 1964; Davis et al., 1983; Kirsten and Mayer, 1967; Vetter and Wittinghofer, 2001). (ii) constitutively active HRas mutant (HRas V12) has been shown to inhibit the differentiation of myoblasts by inhibiting MyoD and Myogenin expression(Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Van Der Burgt et al., 2007). (iii) Downstream signaling of HRas is important for muscle homeostasis as skeletal and cardiac myopathies are observed in individuals carrying constitutively active mutants of HRas(Engler et al., 2021; Konieczny et al., 1989; Lee et al., 2010; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007). HRas is highly regulates by lipid modifications, it undergoes reversible palmitoylation and de-palmitoylation at C-terminal cysteines, which regulate the intracellular distribution and activity of HRas(Gutierrez et al., 1989; Lu and Hofmann, 1995; Rocks et al., 2005). In this study, we show that 15d-PGJ2 is synthesized and released by senescent myoblasts upon treatment with Doxo. 15d-PGJ2, taken up by the myoblasts, covalently modifies HRas at cysteine 184 and activates it. We also show that previously reported inhibition of differentiation of myoblasts by 15d-PGJ2 depends on HRas C-terminal cysteines, notably cysteine 184. This study provides a mechanism by which prostaglandins secreted as SASP inhibit the differentiation of myoblasts, affecting muscle homeostasis in patients undergoing chemotherapy.

Results

Doxorubicin (Doxo) treatment induces senescence in mouse skeletal muscles and C2C12 mouse myoblasts

Doxorubicin-mediated DNA damage has been shown to induce senescence in cells(Di Leonardo et al., 1994; Hu and Zhang, 2019; Robles and Adami, 1998). Therefore, we injected B6J mice with Doxorubicin (Doxo) (5mg/kg) every 3 days for 9 days and observed induction of DNA damage-mediated senescence in hindlimb skeletal muscles (Fig. S1A). We observed an increase in the expression of p21 and increased nuclear levels of the DNA damage marker γH2A.X in mouse Gastrocnemius muscles (Fig. 1A and B). We also observed a significant increase in the mRNA levels of known senescence markers (p16 and p21), SASP factors (CXCL1, CXCL2, TNF1α, IL6, TGFβ1) in skeletal muscles of mice treated with Doxo compared to that of mice treated with saline (Fig. 1C). These observations suggest that there is induction of senescence in skeletal muscles of mice upon treatment with Doxo.

Prostaglandin synthesis and release by Doxo-induced senescent cells inhibits myoblast differentiation.

A. Expression and localization of tumor suppressor protein p21, measured by immunofluorescence, in hindlimb skeletal muscles of mice after 11 days of treatment with Doxo (5 mg/kg) or Saline.

B. Representative confocal micrograph of expression of γH2A.X in the gastrocnemius muscle of mice treated with Doxo (5 mg/kg) or Saline.

C. Expression of mRNAs of senescence markers (p16 and p21), SASP factors (CXCL1, CXCL2, TNF1α, IL6, TGFβ1), and enzymes involved in the biosynthesis of prostaglandin PGD2/15d-PGJ2 (PTGS1, PTGS2, PTGDS), measured by qPCR, in hindlimb skeletal muscles of mice after 11 days of treatment with Doxo (5 mg/kg) or Saline.

D. A representative confocal micrograph and a scatter plot of the nuclear area of C2C12 myoblasts, measured by immunofluorescence, after 16 days of treatment with Doxo (150 nM) or DMSO.

E. Expression of mRNAs of prostaglandin biosynthetic enzymes, measured by qPCR, in C2C12 myoblasts after treatment with Doxo (150 nM) or DMSO.

F. Concentration of 15d-PGJ2 released from quiescent or senescent C2C12 cells. (Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

We then measured the induction of senescence in vitro in C2C12 mouse myoblasts after treatment with Doxo (Fig. S1B). We treated C2C12 myoblasts with Doxo (150 nM) and observed a flattened morphology and increased cell size (Fig. S1C). We also observed a significant increase in the size of nuclei in C2C12 cells after treatment with Doxo (Fig. 1D) and increased senescence-associated β galactosidase (SA-β gal) activity in cells treated with Doxo (Fig. S1D). These observations suggest the induction of senescence in C2C12 cells after treatment with Doxo.

Doxo-mediated senescence induces synthesis and release of 15d-PGJ2 in C2C12 myoblasts and mouse skeletal muscle

Synthesis of prostaglandins by senescent cells has previously been reported(Wiley et al., 2021; Wiley and Campisi, 2021). Specifically, levels of PGD2 and its metabolite 15d-PGJ2 have been shown to be significantly increased. Therefore, we measured the levels of mRNA of enzymes involved in the synthesis of PGD2/15d-PGJ2 (PTGS1, PTGS2, and PTGDS), in the gastrocnemius muscle of mice after treatment with Doxo. We observed a significant increase in the mRNA levels of PTGS1, PTGS2, and PTGDS enzymes in the skeletal muscle of mice treated with Doxo (Fig. 1C). We also observed a time-dependent increase in the mRNA levels of PTGS1, PTGS2, PTGDS, and PTGES enzymes in C2C12 cells treated with Doxo compared to Day 0 (Fig. 1E). Expression of enzyme PTGES was elevated on Day 4, whereas the expression of Prostaglandin D synthase (PTGDS) increased only after Day 8, reaching maximum expression on Day 12. These observations suggest an increase in the synthesis of prostaglandins in senescent cells.

15d-PGJ2 is a non-enzymatic dehydration product of PGD2(Shibata et al., 2002). We observed an increase in the mRNA levels synthetic enzymes of 15d-PGJ2 in senescent C2C12 cells. Therefore, we measured the levels of extracellular 15d-PGJ2. We collected the conditioned medium from senescent and quiescent C2C12 cells and observed an increase in the concentration of 15d-PGJ2 in the medium of senescent cells (Fig. 1F). This shows an increased release of 15d-PGJ2 by senescent myoblasts.

Prostaglandin PGD2 and its metabolites in the conditioned medium of senescent cells inhibit the differentiation of C2C12 myoblasts

15d-PGJ2 (the final non-enzymatic dehydration product of PGD2) has been shown to inhibit the differentiation of myoblasts(Hunter et al., 2001). We observed the release of 15d-PGJ2 by senescent cells, showing that 15d-PGJ2 is a SASP factor (Fig. 1F). Conditioned medium of senescent cells inhibits the differentiation of myoblasts in myotonic dystrophy type 1(Conte et al., 2023). Therefore, we tested whether 15d-PGJ2, the terminal dehydration product of PGD2, is required for the inhibitory effect of SASP on the differentiation of myoblasts. We treated C2C12 myoblasts with the conditioned medium of senescent cells or senescent cells treated with 30 µM of AT-56 (a well-characterized inhibitor of prostaglandin D synthase (PTGDS))(Hu et al., 2021; S. Hu et al., 2023; Shunfeng Hu et al., 2023; Irikura et al., 2009). We observed a decrease in the fusion index of the C2C12 myoblasts treated with the conditioned medium of senescent cells (Fig. 2A), suggesting that SASP factors decrease the differentiation of myoblasts. This decrease in inhibition was rescued in myoblasts treated with the conditioned medium of senescent cells treated with AT-56 (Fig. 2A). This suggests that prostaglandins PGD2/15d-PGJ2 released by senescent cells as SASP factors can inhibit the differentiation of myoblasts.

15d-PGJ2 inhibits differentiation of myoblasts.

A. Expression of MHC protein and the fusion of myoblasts in myotubes, measured by immunofluorescence, after treatment with conditioned medium of senescent cells treated with PTGDS inhibitor AT-56 (30 µM) or DMSO

B. Normalized number of C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO

C. Expression of mRNAs of markers of differentiation (MyoD, MyoG, and MHC), measured by qPCR, in C2C12 myoblasts treated with 15d-PGJ2 (1 µM, 2 µM, or 4 µM) or DMSO.

D. Expression of MHC protein and the fusion of myoblasts in syncytial myotubes, measured by immunofluorescence, after treatment with 15d-PGJ2 (4 µM) or DMSO.

E. Expression of MHC protein, measured by immunoblotting, in primary human skeletal myoblasts after treatment with 15d-PGJ2 (1 µM, 2 µM, or 4 µM) or DMSO for 5 days.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 inhibits the proliferation and differentiation of mouse and human myoblasts

15d-PGJ2 has been shown to affect the proliferation of cancer cell lines, both positively and negatively (Chen et al., 2003; Choi et al., 2020; Slanovc et al., 2024; Yen et al., 2014). We measured the effect of 15d-PGJ2 on the proliferation of C2C12 myoblasts. We treated C2C12 myoblasts with 15d-PGJ2 (10 µM) or DMSO for 72 hours. We observed a decrease in the proliferation of C2C12 cells after treatment with 15d-PGJ2 (Fig. 2B). The doubling time of C2C12 cells was also increased upon treatment with 15d-PGJ2 (57.24 hours) compared to DMSO (13.76 hours). This suggests that 15d-PGJ2 decreases the proliferation of C2C12 myoblasts.

We measured the differentiation of C2C12 mouse and primary human myoblasts after treatment with 15d-PGJ2. C2C12 myoblasts showed cell death after 24 hours and 48 hours upon treatment with 15d-PGJ2 (10 µM and 5 µM) respectively in the C2C12 differentiation medium (Fig. S2A). Therefore, we treated differentiating myoblasts with 15d-PGJ2 (1 µM, 2 µM, and 4 µM) for 5 days. We observed a dose-dependent decrease in the mRNA levels of MyoD, MyoG, and MHC in differentiating C2C12 cells after treatment with 15dPGJ2 (Fig. 2C). We also observed a decrease in the fusion of myoblasts into myotubes in C2C12 cells treated with 15d-PGJ2 (4 µM) compared to DMSO (Fig. 2D). We also observed a dose-dependent decrease in the protein levels of MHC in differentiating primary human myoblasts upon treatment with 15d-PGJ2 (Fig. 2E). These observations suggest that 15d-PGJ2 inhibits the differentiation of both mouse and human myoblasts.

Biotinylated 15d-PGJ2 covalently modifies HRas at Cysteine 184

15d-PGJ2 has been shown to covalently modify several proteins including p53 and NF-κB, which are involved in several key biological processes(Marcone and Fitzgerald, 2013). HRas was identified to be covalently modified by 15d-PGJ2 at cysteine 184 in NIH3T3 and Cos1 cells(Luis Oliva et al., 2003). Therefore, we tested whether 15d-PGJ2 could covalently modify HRas in C2C12 cells. We treated C2C12 cells expressing the EGFP-tagged wild-type HRas with biotinylated 15d-PGJ2 (5 µM). We then immunoprecipitated biotinylated 15d-PGJ2 using streptavidin. We observed a significant increase in the pulldown of HRas upon treatment with 15d-PGJ2-biotin compared to DMSO (Fig. 3A), suggesting an interaction between 15d-PGJ2 and HRas. To measure the role of individual C-terminal cysteines in the binding of HRas with 15d-PGJ2, we treated C2C12 cells expressing the EGFP-tagged C181S and C184S mutants of HRas with biotinylated 15d-PGJ2 (5 µM), and immunoprecipitated using streptavidin. We observed that the intensity of EGFP-tagged HRas was significantly decreased in cells expressing the C184S mutant but not in those expressing the C181S mutant (Fig. 3A). This suggests that 15d-PGJ2 covalently modifies HRas at cysteine 184 in C2C12 cells.

15d-PGJ2 covalently modifies HRas at Cysteine 184 and activates the HRas-MAPK pathway via the electrophilic cyclopentenone ring.

A. Streptavidin-immunoprecipitation of EGFP-HRas, measured by immunoblotting, in C2C12 cells after 3 hours of treatment with 15d-PGJ2-Biotin (5 µM).

B. Representative confocal micrograph of Fluorescence Resonance Energy Transfer (FRET) between EGFP-tagged HRas (EGFP-HRas) and mCherry-tagged Ras binding domain (RBD) of RAF kinase (mCherry-RAF RBD).

C. Activation of the EGFP-tagged wild type HRas (HRas WT), measured by FRET, before and after 1 hour of treatment with 15d-PGJ2 (10 µM) after starvation for 24 hours.

D. Activation of the EGFP-tagged C-terminal cysteine mutants of HRas (HRas C181S and HRas C184S), measured by FRET, before and after 1 hour of treatment with 15d-PGJ2 (10 µM) after starvation for 24 hours.

E. Phosphorylation of Erk (42 kDa and 44 kDa), measured by immunoblotting, in C2C12 cells after 1 hour of treatment with 15d-PGJ2 (5 µM, 10 µM) or DMSO after starvation for 24 hours.

F. Phosphorylation of Erk (42 kDa and 44 kDa), measured by immunoblotting, in C2C12 cells after 1 hour of treatment with 15d-PGJ2 (10 µM)/ 9,10-dihydro-15d-PGJ2 (10 µM) or DMSO after starvation for 24 hours.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 increases the FRET between EGFP-HRas and mCherry-RAF RBD in wild-type and C181S mutant but not in the C184S mutant of HRas

We next tested the effect of covalent modification of HRas by 15d-PGJ2 on HRas GTPase activity using FRET. mCherry RAF RBD is a well characterized sensor of activity of HRas. RAF RBD binds to the activated HRas upon activation of HRas, allowing FRET between EGFP and mCherry(Rocks et al., 2005). We co-expressed EGFP-tagged HRas (EGFP-HRas) with mCherry-tagged Ras Binding Domain of RAF kinase (mCherry-RAF RBD) in C2C12 myoblasts (Fig. 3B). We measured the efficiency of FRET between EGFP and mCherry using an ImageJ plugin, FRET analyzer(Hachet-Haas et al., 2006). We compared the mean acceptor normalized FRET index in C2C12 myoblasts expressing EGFP-HRas WT before and after treatment of 15d-PGJ2 (10 µM) for 1 hour. We observed an increase in the mean acceptor normalized FRET index upon treatment with 15d-PGJ2 (Fig. 3C). This suggests that 15d-PGJ2 activates HRas. To measure the role of individual C-terminal cysteines in 15d-PGJ2 mediated activation of HRas, we co-expressed EGFP-HRas C181S or C184S with mCherry-RAF RBD in C2C12 myoblasts. We measured the mean acceptor normalized FRET index before and after 1 hour of treatment with 15d-PGJ2 (10 µM). We observed an increase in the mean acceptor normalized FRET index in cells expressing EGFP-HRas C181S upon treatment with 15d-PGJ2 but not in cells expressing EGFP-HRas C184S (Fig. 3D). These observations suggest that activation of HRas by 15d-PGJ2 occurs in a cysteine 184 dependent manner.

15d-PGJ2 increases phosphorylation of Erk (Thr202/Tyr204) but not Akt (S473) in C2C12 myoblasts

HRas regulates two major downstream signaling pathways, the MAP kinase (MAPK) pathway and the PI3 kinase (PI3K) pathway(Pylayeva-Gupta et al., 2011). We tested the effects of treatment with 15d-PGJ2 on these two downstream signaling pathways by measuring the phosphorylation of Erk (42 kDa and 44 kDa) and Akt proteins in C2C12 cells. We treated C2C12 cells with 15d-PGJ2 (5 µM and 10 µM) or DMSO for 1 hr (after 24 hrs. of serum starvation) and observed a dose-dependent increase in the phosphorylation of Erk (T202/Y204) (42 kDa) but not of Erk (44 kDa) (Fig. 3E). We did not observe an increase in the phosphorylation of Akt (S473) in C2C12 cells after treatment with 15d-PGJ2 (Fig. S3C). These observations suggest that 15d-PGJ2 activates the MAPK signaling pathway, but not the PI3K signaling pathway.

15d-PGJ2 contains a reactive electrophilic center in its cyclopentenone ring, that can react with cysteine residues of proteins(Luis Oliva et al., 2003). We tested its role in activating the MAPK signaling pathway. We measured the phosphorylation of Erk (42kDa and 44 kDa) in C2C12 cells after treatment with cells with 9,10-dihydro-15d-PGJ2 (10 µM), a 15d-PGJ2 analog which is devoid of the electrophilic center, for 1 hr (after 24 hr. of serum starvation). We observed that the phosphorylation of Erk (42 kDa and 44 kDa) in C2C12 cells treated with 9,10-dihydro-15d-PGJ2 was significantly reduced as compared to the treatment with 15d-PGJ2 (Fig. 3F). This shows that 15d-PGJ2 activates the HRas-MAPK signaling pathway via the electrophilic center in its cyclopentenone ring.

15d-PGJ2 increases the localization of EGFP-tagged HRas at the plasma membrane compared to the Golgi in a C-terminal cysteine-dependent manner

15d-PGJ2 covalently modifies cysteine 184 and activates HRas signaling (Fig. 3). Reversible palmitoylation of cysteine 181 and cysteine 184 in the C-terminal tail of HRas regulate intracellular distribution and signaling of HRas. Inhibition of palmitoylation of the C-terminal cysteine 181, either by a palmitoylation inhibitor 2-Bromopalmitate or by mutation to serine, causes accumulation of HRas at the Golgi compared to the plasma membrane and alters activity(Rocks et al., 2005). Therefore, we tested whether the modification of 15d-PGJ2 alters the intracellular distribution of HRas. We co-expressed the EGFP-tagged wild type and the C181S and the C184S mutants of HRas with a previously reported marker of Golgi(Shaner et al., 2008) in C2C12 cells and stained the cells with plasma membrane marker WGA-633 (Fig. 4A). We compared Rmean, the ratio of mean EGFP-HRas intensity at the Golgi to the mean HRas intensity at the plasma membrane, to measure the distribution of HRas between the plasma membrane and the Golgi. We measured the intracellular distribution of HRas between the Golgi and the plasma membrane in C2C12 cells after treatment with 15d-PGJ2 (10 µM) for 24 hours and observed a decrease in the Rmean of C2C12 cells expressing the wild-type HRas after treatment with 15d-PGJ2 (Fig. 4B). However, we did not observe a change in the Rmean of C2C12 cells expressing HRas C181S or HRas C184S after treatment with 15d-PGJ2 (Fig. 4C). These observations suggest that 15d-PGJ2 increases the localization of HRas at the plasma membrane as compared to that in the Golgi in an HRas C-terminal cysteine-dependent manner.

15d-PGJ2 controls the intracellular distribution of HRas and differentiation of C2C12 cells in an HRas C-terminal cysteine-dependent manner.

A. Representative confocal micrograph of C2C12 myoblasts showing localization of EGFP-tagged HRas between the plasma membrane (stained with Alexa Fluor 633 conjugated Wheat Germ Agglutinin) and the Golgi (labelled with TagRFP-tagged Golgi resident GalT protein). A statistic Rmean was defined as the ratio of mean HRas intensity at the Golgi to the mean HRas intensity at the plasma membrane.

B. Distribution of the wild-type HRas between the Golgi and the plasma membrane, measured by Rmean, in C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO for 24 hours.

C. Distribution of the C-terminal cysteine mutants of HRas between the Golgi and the plasma membrane, measured by Rmean, in C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO for 24 hours.

D. Expression of mRNAs of known markers of differentiation (MyoD, MyoG, and MHC), measured by qPCR, in differentiating C2C12 myoblasts expressing the EGFP-tagged wild-type and the C-terminal cysteine mutants of HRas after treatment with 15d-PGJ2 (4 µM) or DMSO for 5 days.

E. Expression of MHC protein, measured by immunoblotting, in differentiating C2C12 myoblasts expressing the EGFP-tagged wild-type and the C-terminal cysteine mutants of HRas after treatment with 15d-PGJ2 (4 µM) or DMSO for 5 days.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 mediated inhibition of differentiation of C2C12 cells is rescued by C181S and C184S mutants of HRas

HRas inhibits the differentiation of C2C12 myoblasts(Engler et al., 2021; Konieczny et al., 1989; Lassar et al., 1989; Lee et al., 2010; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007).15d-PGJ2 covalently modifies cysteine 184 and activates HRas (Fig. 3). Therefore, we tested whether the inhibition of myoblast differentiation by 15d-PGJ2 depends on the activation of HRas signaling by modification of the C-terminal cysteine 184. We expressed the wild-type and the cysteine mutants of HRas (EGFP-HRas WT/C181S/C184S) in C2C12 myoblasts and treated the cells with 15d-PGJ2 (4 µM) or DMSO during differentiation. We observed a decrease in the levels of mRNA of MHC in C2C12 cells expressing HRas WT and HRas C181S after 5 days of treatment with 15d-PGJ2. We did not observe this in expressing HRas C184S (Fig. 4D). We also observed a significant decrease in the protein levels of MHC in differentiating C2C12 cells expressing HRas WT and HRas C181S after treatment with 15d-PGJ2 (Fig. 4E). This decrease was partially rescued in cells expressing HRas C184S (Fig. 4E). These observations suggest that the inhibition of myoblast differentiation by 15d-PGJ2 depends on modification of HRas C-terminal cysteine 184.

Discussion

Senescence is characterized by an irreversible arrest in cell proliferation(Hayflick, 1965). Cells undergo senescence because of a myriad of stresses, including DNA damage, mitochondrial damage, and oncogene overexpression(Bihani et al., 2007, 2004; Casar et al., 2018; Chen and Ames, 1994; Chen et al., 1998; Coppé et al., 2008; D’Adda Di Fagagna, 2008; D’Adda Di Fagagna et al., 2003; Di Leonardo et al., 1994; Franza et al., 1986; Land et al., 1983; Robles and Adami, 1998; Serrano et al., 1997; Wiley et al., 2016; Woods et al., 1997). Senescent cells exhibit a multi-faceted physiological response, where they exhibit a flattened morphology, increase in cell size(Chen and Ames, 1994; Serrano et al., 1997), upregulation of tumor suppressor proteins(Calabrese et al., 2009; Lowe et al., 2004; Stein et al., 1990; Zindy et al., 2003), expression of neutral pH active β-galactosidase(Dimri et al., 1995; Lee et al., 2006), and altered metabolic state(Bittles and Harper, 1984; Jones et al., 2005; Wiley and Campisi, 2021, 2016; Zwerschke et al., 2003). Arachidonic acid metabolism is upregulated in senescent cells, which leads to increased synthesis of eicosanoid prostaglandins, which regulate the physiology of senescent cells(Wiley et al., 2021; Wiley and Campisi, 2021). Senescent cells exhibit a secretory phenotype (SASP) consisting of a variety of bioactive molecules including cytokines and chemokines, growth factors, matrix metalloproteases etc(Coppé et al., 2008). Senescent cells influence the surrounding cells via the SASP factors, which regulate proliferation, migration, and other cell biological processes in the neighboring cells(Campisi, 2005). SASP-mediated perturbations in the microenvironment are implicated in several senescence-associate pathologies(Wiley and Campisi, 2021). Senescent fibroblasts increase the proliferation of premalignant and malignant epithelial cells(Krtolica et al., 2001). Conditioned medium of senescent fibroblasts promoted tumorigenesis in mouse keratinocytes(Dilley et al., 2003). Senescent fibroblasts transform pre-malignant breast cancer cells into invasive, tumor-forming cells(Parrinello et al., 2005). Senescence in muscle stem cells induces sarcopenia via activation of p38 MAP kinase pathway and transient inhibition of the p38 MAP kinases rejuvenates aged muscle stem cells to ameliorate sarcopenia(Cosgrove et al., 2014). Senescent cells inhibit the differentiation of myoblasts by secretion of IL6 by senescent muscle stem cells in myotonic dystrophy(Conte et al., 2023).

In this study, we show that senescent myoblasts synthesize and release eicosanoid prostaglandin 15-deoxy-Δ12,14-prostaglandin J2 (15d-PGJ2) (Fig. 1E and F), the terminal non-enzymatic dehydration product of prostaglandin PGD2(Shibata et al., 2002). We used Doxorubicin (Doxo) to induce senescence in C2C12 myoblasts and showed that the conditioned medium of senescent C2C12 cells inhibits differentiation of C2C12 myoblasts (Fig. 2A). Inhibition of synthesis of PGD2 by treatment of senescent cells with AT-56, a well-characterized inhibitor of prostaglandin D synthase(Hu et al., 2021; S. Hu et al., 2023; Shunfeng Hu et al., 2023; Irikura et al., 2009), rescued this inhibitory effect of the conditioned medium (Fig. 2A). A study has shown that prostaglandin PGD2 inhibits differentiation of C2C12 myoblasts(Veliça et al., 2010), but the authors noted that knockout of DP1 and DP2 (the known receptors of prostaglandins PGD2) does not abrogate inhibition of differentiation of myoblasts by PGD2. This observation suggested that PGD2 might inhibit the differentiation of myoblasts by a receptor-independent mechanism, possibly by its spontaneous non-enzymatic dehydration to 15d-PGJ2. 15d-PGJ2 has been suggested to be an endogenous ligand of PPARγ(Li et al., 2019). However, the inhibition of PPARγ did not abrogate the inhibition of differentiation of C2C12 myoblasts by 15d-PGJ2, suggesting existence other possible mechanisms(Hunter et al., 2001). 15d-PGJ2 has varied effects on cell physiology in a context dependent manner. On one hand, 15d-PGJ2 promotes tumorigenesis by inducing epithelial to mesenchymal transition in breast cancer cell line MCF7(Choi et al., 2020), 15d-PGJ2 inhibits proliferation of A549, H1299, and H23 lung adenocarcinoma cells via induction of ROS and activation of apoptosis(Slanovc et al., 2024). Here, we show that 15d-PGJ2 inhibits the proliferation and the differentiation of C2C12 myoblasts (Fig. 2B, C and D).

15d-PGJ2 contains an electrophilic cyclopentenone ring in its structure, allowing 15d-PGJ2 to covalently modify and form a Michael adducts with cysteine residues of proteins(Shibata et al., 2002). A previous proteomic study in endothelial cells showed biotinylated 15d-PGJ2 covalently modified over 300 proteins, which regulate several physiological processes including cell cycle (MAPK1, MCM4), cell metabolism (Fatty acid synthase, Isocitrate dehydrogenase), apoptosis (PDCD6I), translation regulation (Elongation factor 1 and 2, EIF4A-I), intracellular transport (Importin subunit β1, Exportin 2, Kinesin 1 heavy chain)(Marcone and Fitzgerald, 2013). Another proteomic study in neuronal cells suggested that 15d-PGJ2 modifies several proteins including chaperone HSP8A, glycolytic proteins Enolase 1 and 2, GAPDH, PKM1, cytoskeleton proteins Tubulin β2b, β actin, GFAP etc(Yamamoto et al., 2011). This study also showed modification of a peptide fragments homologous to IκB kinase β, Thioredoxin, and a small molecule GTPase HRas. 15d-PGJ2 modifies HRas in NIH3T3 and Cos1 cells(Luis Oliva et al., 2003) and IMR90 cells(Wiley et al., 2021). Modification by 15d-PGJ2 led to activation of HRas, judged by an increase in GTP bound HRas. It is clear that 15d-PGJ2 is capable of modifying numerous proteins in different contexts. In spite of these observations, the functional relevance of these modifications in numerous contexts remains to be mapped. Here we focused on the role of 15d-PGJ2 in the context of senescence and skeletal muscle differentiation. In this study, we showed that 15d-PGJ2 covalently modifies HRas at cysteine 184 but not cysteine 181 in C2C12 myoblasts (Fig. 3A). We showed by FRET microscopy that modification of HRas by 15d-PGJ2 in HRas WT and HRas C181S activates HRas in C2C12 cells, but 15d-PGJ2 is unable to activate HRas C184S in this context (Fig. 3C and D). This observation shows a direct link between modification of HRas by 15d-PGJ2 and activation of HRas GTPase.

HRas activates two major downstream signaling pathways, the HRas-MAPK and the HRas-PI3K pathway(Pylayeva-Gupta et al., 2011). We showed that covalent modification of HRas by 15d-PGJ2 via the electrophilic cyclopentenone ring activates HRas (Fig. 3C and D) and activates the HRas-MAPK pathway, demonstrated by an increase in the phosphorylation of Erk after treatment with 15d-PGJ2 (Fig. 3E and However, we did not observe activation of the HRas-PI3K pathway, as we did not see an increase in the phosphorylation of Akt after treatment with 15d-PGJ2 (Fig. S3C). MAPK and PI3K pathways are known regulators of muscle differentiation(Bennett and Tonks, 1997; Rommel et al., 1999), where inhibition of the RAF-MEK-Erk pathway or activation of PI3K pathway promotes differentiation of myoblasts. Preferential activation of HRas-MAPK over HRas-PI3K pathway after treatment with 15d-PGJ2 can be a possible mechanism by which 15d-PGJ2 can inhibit differentiation of myoblasts. HRas is known to regulate differentiation of myoblasts in different contexts. Constitutively active HRas signaling by oncogenic HRas mutant (HRas V12) leads to inhibition of differentiation of myoblasts(Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Van Der Burgt et al., 2007). Here we showed that inhibition of differentiation of myoblasts after 15d-PGJ2 is partially rescued in cells expressing C184S mutants of HRas but not the wild type or C181S mutant (Fig. 4D and E and S4E). HRas C184S did not get modified by 15d-PGJ2 (Fig. 3A). These observations suggest that the inhibition of differentiation of myoblasts by 15d-PGJ2 is partially dependent on modification of HRas by 15d-PGJ2.

Cysteine 181 and 184 in the C-terminal of HRas regulate the intracellular distribution of HRas between the plasma membrane and the Golgi by reversible palmitoylation and depalmitoylation(Rocks et al., 2005). Inhibition of palmitoylation of C-terminal cysteine 181, wither by treatment with protein palmitoylation inhibitor 2-bromopalmitate or mutation of cysteine to serine, leads to accumulation of HRas at the Golgi. Intracellular localization of HRas maintains two distinct pools of HRas activity, where the plasma membrane pool shows a faster activation followed by short kinetics and the Golgi pool shows a slower activation but a sustained activation(Agudo-Ibáñez et al., 2015; Busquets-Hernández and Triola, 2021; Lorentzen et al., 2010; Rocks et al., 2005). We showed that modification of HRas by 15d-PGJ2 alters the intracellular distribution of HRas. We showed that modification of HRas by 15d-PGJ2 leads to an increase in the localization of the wild type HRas at the plasma membrane compared to the Golgi (Fig. 4B). We did not observe any changes in the intracellular distribution of HRas C181S or HRas C184S after treatment with 15d-PGJ2 (Fig. 4C). HRas C184S is not modified by 15d-PGJ2, but HRas C181S is modified by 15d-PGJ2 (Fig. 3A). This suggests that intracellular redistribution of HRas due to modification by 15d-PGJ2 at cysteine 184 requires palmitoylation of cysteine 181.

Previous reports suggest that downstream signaling of HRas depends on the intracellular localization of HRas(Rocks et al., 2005; Santra et al., 2019). For example, targeted localization of HRas at the ER membrane induced expression of cell-migration genes. Localization of HRas at the plasma membrane showed strong correlation with expression of cell cycle genes, particularly the MAPK signaling pathway. Localization of HRas at the plasma membrane also showed negative correlation with genes associated with the PI3K-Akt pathway. Here we showed that the intracellular distribution of HRas regulates differentiation of myoblasts. in order to show this, we used the constitutively active mutant of HRas (HRas V12) which has been shown to inhibit differentiation of myoblasts(Engler et al., 2021; Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007). We expressed cysteine mutants of HRas V12 in C2C12 myoblasts and found that HRas V12 C181S localized predominantly at the Golgi whereas HRas V12 and HRas V12 C184S localized at both the plasma membrane and the Golgi (Fig. S4A). When differentiated, we observed that C2C12 cells expressing HRas V12 C181S differentiated but HRas V12 or HRas V12 C184S did not differentiate (Fig. S4B, C, and D). These observations suggest alteration of intracellular distribution of HRas affects HRas mediated inhibition of differentiation of myoblasts.

Doxorubicin (Doxo) is a widely used chemotherapy agent for treatment of cancers(Johnson-Arbor and Dubey, 2022). Treatment with Doxo induces senescence. Doxo mediated DNA damage leads to p53, p16, and p21 dependent senescence in human fibroblasts(Di Leonardo et al., 1994; Robles and Adami, 1998). On the other hand, treatment with doxorubicin leads to decrease in muscle mass and cross-sectional area, leading to chemotherapy induced cachexia(Hiensch et al., 2020). Several mechanisms have been proposed behind chemotherapy induced cachexia, including generation of reactive oxygen species(Gilliam and St. Clair, 2011), activation of proteases like calpain and caspases(Gilliam et al., 2012; Smuder et al., 2011), and impaired insulin signaling(de Lima Junior et al., 2016). This study provides a possible mechanism behind chemotherapy induced loss of muscle mass and functioning. Induction of senescence in myoblasts by treatment with Doxo could lead to increased synthesis and release of 15d-PGJ2 by senescent cells which could be taken up by myoblasts in the microenvironment. The lipid could covalently modify and activates HRas at cysteine 184 to inhibit differentiation of myoblasts. Therefore, targeting the synthesis and release of 15d-PGJ2 by senescent cells could serve as an important target to promote skeletal muscle homeostasis in cancer patients.

eLife Assessment

This manuscript provides useful information about the lipid metabolite 15d-PGJ2 as a potential regulator of myoblast senescence. The authors provide experimental evidence that 15d-PGJ2 inhibits myoblast proliferation and differentiation by binding and regulating HRas. However, the manuscript is incomplete in its current form, as it lacks robust support from the data regarding the main conclusions related to senescence and technical concerns related to the senescence models used in this study.

Reviewer #1 (Public Review):

Summary:

The authors show that upon treatment with Doxorubicin (Doxo), there is an increase in senescence and inflammatory markers in the muscles. They also show these genes get upregulated in C2C12 myoblasts when treated with conditioned media or 15d-PGJ2. 15dPGJ2 induces cell death in the myoblasts, decreases proliferation (measured by cell numbers), and decreases differentiation and fusion. 15d-PGJ2 modified Cys184 of HRas, which is required for its activation as indicated by the FRET analysis with RAF RBD. They also showed that 15d-PGJ2 activates ERK signaling, but not Akt signaling, through the electrophilic center. 15d-PGJ2 inhibits Golgi localization of HRAS (only WT, not C181 or C184 mutant). They also showed that expressing the WT HRas followed by 15d-PGJ2 treatment led to a decrease in the levels of MHC mRNA and protein, and this defect is dependent on C184. This is a well-written manuscript with interesting insights into the mechanism of action of 15d-PGJ2. However, some clarification and experiments will help the paper advance the field significantly.

Strengths:

The data clearly shows that 15d-PGJ2 has a negative role in the myoblast cells and that it leads to modification of HRas protein. Moreover, the induction of biosynthetic enzymes in the PGD2 pathway also supports the induction of 15d-PGJ2 in Doxorubicin-treated cells. Both conditioned media experiments and the 15d-PGJ2 experiments show that 15d-PGJ2 could be the active component secreted by the senescent myoblasts.

Weaknesses:

The genes that are upregulated in the muscles upon injection with Doxo are also markers for inflammation. Since Doxo is also known to induce systemic inflammation, it is important to delineate these two effects (inflammatory cells vs senescent cells). The expression of beta Gal and other markers of senescence in the tissue sections will help to delineate these.

In Figure 2, where the defect in the differentiation of myoblasts upon treatment with 15d-PGJ2 is shown, most of the cells die within 48 hours at higher concentrations, making it difficult to perform the experiments. This also shows that 15d-PGJ2 was toxic to these cells. Lower concentrations show a decrease in the differentiation based on the lower number of nuclei in fibers and low expression of MyoD, MyoG, and MHC. However, it is unclear if this is due to increased cell death or defective differentiation. It would be a lot more informative if the cell count, cell division, and cell death could be plotted for these concentrations of the drug during the experiment. Also, in the myoblast experiments, are the effects of treatment with Dox reversible?

In Figure 3, most of the experiments are done at a high concentration, which induces almost complete cell death within 48 hours. Even at such a high concentration of 15dPGJ2, the increase in ERK phosphorylation is minimal.

The experiment Figure 4C shows that C181 and C84 mutants of the HRas show higher levels in Golgi compared with WT. However, this could very well be due to the defect in palmitoylation rather than the modification with 15d-PGJ2. Though the authors allude to the possibility that intracellular redistribution of HRas by 15d-PGJ2 requires C181 palmitoylation, the direct influence of C184 modification on C181 palmitoylation is not shown. To have a meaningful conclusion, the authors need to compare the palmitoylation and modification with 15d-PGJ2.

To test if the inhibition of myoblast differentiation depends on HRas, they overexpressed the HRas and mutants in the C2C12 lines. However, this experiment does not take the endogenous HRAs into consideration, especially when interpreting the C184 mutant. An appropriate experiment to test this would be to knock down or knock out HRas (or make knock-in mutations of C184) and show that the effect of 15d-PGJ2 disappears. Moreover, in this specific experiment, it is difficult to interpret without a control with no HRas construct and another without the 15d-PGJ2 treatment.

Moreover, the overall study does not delineate the toxic effects of 15d-PGJ2 from its effect on the differentiation.

Reviewer #2 (Public Review):

Summary:

In this study, Swarang and colleagues identified the lipid metabolite 15d-PGJ2 as a potential component of senescent myoblasts. They proposed that 15d-PGJ2 inhibits myoblast proliferation and differentiation by binding and regulating HRas, suggesting its potential as a target for restoring muscle homeostasis post-chemotherapy.

Strengths:

The regulation of HRas by 15d-PGJ2 is well controlled.

Weaknesses:

The novelty of the study is compromised as the activation of PGD and 15d-PGJ2, as well as the regulation of HRas and cell proliferation, have been previously reported. Additionally, there are major technical concerns related to the senescence models, limiting data interpretation regarding the relevance to senescent cells.

Major concerns:

(1) The C2C12 cell line is not an ideal model for senescence study due to its immortalized nature and lack of normal p16 expression. A more suitable myoblasts model is recommended, with a more comprehensive characterization of senescence features.

(2) The source of increased PGD or its metabolites in the conditioned medium is unclear. Including other senescence models, such as replicative or oncogene-induced senescence, would strengthen the study. Again, C2C12 is not suitable for replicative senescence due to its immortalized status.

(3) In the in vivo part, it's unclear whether the increased expression of PTGS1, PTGS2, and PTGDS is due to senescence or other side effects of DOXO.

(4) Figure 2A lacks an important control from non-senescent cells during the measurement of C2C12 differentiation in the presence of a conditioned medium. There is no explanation of how differentiation was quantified or how the fusion index was calculated.

Author response:

The following is the authors’ response to the current reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors show that upon treatment with Doxorubicin (Doxo), there is an increase in senescence and inflammatory markers in the muscles. They also show these genes get upregulated in C2C12 myoblasts when treated with conditioned media or 15d-PGJ2. 15dPGJ2 induces cell death in the myoblasts, decreases proliferation (measured by cell numbers), and decreases differentiation and fusion. 15d-PGJ2 modified Cys184 of HRas, which is required for its activation as indicated by the FRET analysis with RAF RBD. They also showed that 15d-PGJ2 activates ERK signaling, but not Akt signaling, through the electrophilic center. 15d-PGJ2 inhibits Golgi localization of HRAS (only WT, not C181 or C184 mutant). They also showed that expressing the WT HRas followed by 15d-PGJ2 treatment led to a decrease in the levels of MHC mRNA and protein, and this defect is dependent on C184. This is a well-written manuscript with interesting insights into the mechanism of action of 15d-PGJ2. However, some clarification and experiments will help the paper advance the field significantly.

Strengths:

The data clearly shows that 15d-PGJ2 has a negative role in the myoblast cells and that it leads to modification of HRas protein. Moreover, the induction of biosynthetic enzymes in the PGD2 pathway also supports the induction of 15d-PGJ2 in Doxorubicin-treated cells. Both conditioned media experiments and the 15d-PGJ2 experiments show that 15d-PGJ2 could be the active component secreted by the senescent myoblasts.

Weaknesses:

The genes that are upregulated in the muscles upon injection with Doxo are also markers for inflammation. Since Doxo is also known to induce systemic inflammation, it is important to delineate these two effects (Inflammatory cells vs senescent cells). The expression of beta Gal and other markers of senescence in the tissue sections will help to delineate these.

As pointed out Doxo induces systemic inflammation along with inducing DNA damage-mediated senescence. Therefore, along with the inflammatory markers of the SASP (CXCL1/2, TNF1α, IL6, PTGS1/2, PTGDS) we also observed an increase in the mRNA levels of canonical markers of DNA damage-mediated senescence. We observed an increase in the mRNA levels of cell cycle and senescence associated proteins p16 and p21 (Fig. 1C). We also observed an increased nuclear accumulation of p21 (Fig. 1A) and increased levels of phosphorylated H2A.X in the nucleus (Fig. 1B).

In Figure 2, where the defect in the differentiation of myoblasts upon treatment with 15d-PGJ2 is shown, most of the cells die within 48 hours at higher concentrations, making it difficult to perform the experiments. This also shows that 15d-PGJ2 was toxic to these cells. Lower concentrations show a decrease in the differentiation based on the lower number of nuclei in fibers and low expression of MyoD, MyoG, and MHC. However, it is unclear if this is due to increased cell death or defective differentiation. It would be a lot more informative if the cell count, cell division, and cell death could be plotted for these concentrations of the drug during the experiment.

We measured the viability of C2C12 cells after 24 hours of treatment with 15d-PGJ2 using the MTT assay and observed that the viability of cells was decreased after treatment with 15d-PGJ2 (10 µM) but not with 15d-PGJ2 (1 µM, 2 µM, 4 µM, or 5 µM) (see Fig. S2A of the updated manuscript). The results and figures of the manuscript have been updated accordingly.

Also, in the myoblast experiments, are the effects of treatment with Dox reversible?

The treatment with Doxorubicin is irreversible as the senescent phenotype was not reversed after withdrawal of Doxorubicin, even after 20 days.

In Figure 3, most of the experiments are done at a high concentration, which induces almost complete cell death within 48 hours.

Figure 3 is an acute experiment for only 1 hour, at which time no cell death was observed. Specifically, we measured the phosphorylation of Erk and Akt proteins after 1 hour of treatment with 15d-PGJ2 (10 µM) during which we did not observe any cell death.

Even at such a high concentration of 15dPGJ2, the increase in ERK phosphorylation is minimal.

We observe a ~30% increase in the phosphorylation of Erk proteins after treatment with 15d-PGJ­2 in 0.2% serum medium compared to treatment with vehicle (DMSO). This is reproducible and significant.

The experiment Figure 4C shows that C181 and C84 mutants of the HRas show higher levels in Golgi compared with WT. However, this could very well be due to the defect in palmitoylation rather than the modification with 15d-PGJ2.

Our data does not suggest higher levels of C184S mutant in the Golgi compared with WT (Fig. S4A). We observed that the ratio of HRas levels in the Golgi to the HRas levels in the plasma membrane were similar in C2C12 cells expressing HRas C184S and HRas WT (Fig. S4A graph columns 1 and 5).

Though the authors allude to the possibility that intracellular redistribution of HRas by 15d-PGJ2 requires C181 palmitoylation, the direct influence of C184 modification on C181 palmitoylation is not shown. To have a meaningful conclusion, the authors need to compare the palmitoylation and modification with 15d-PGJ2.

Palmitoylation of HRas C181S is required for the localization of HRas at the plasma membrane. The inhibition of palmitoylation of C181, either by mutation (C181S) or treatment with protein palmitoyl transferase inhibitor (2-Bromopalmitate), results in the accumulation of HRas at Golgi(Rocks et al., 2005) (Fig. S4A). Modification of HRas at C184 by 15d-PGJ2 (Fig. 3A) could inhibit the palmitoylation of HRas at C181. However, our data does not support this hypothesis as modification of HRas WT by 15d-PGJ2 does not increase the level of HRas at the Golgi, like in the case of inhibition of cysteine palmitoylation due to C181S mutation.

To test if the inhibition of myoblast differentiation depends on HRas, they overexpressed the HRas and mutants in the C2C12 lines. However, this experiment does not take the endogenous HRAs into consideration, especially when interpreting the C184 mutant. An appropriate experiment to test this would be to knock down or knock out HRas (or make knock-in mutations of C184) and show that the effect of 15d-PGJ2 disappears.

Endogenous HRas (wild type) is present in the C2C12 cells overexpressing the EGFP-tagged HRas constructs. Therefore, we only observe a partial rescue in the differentiation after 15d-PGJ2 treatment in C2C12 cells expressing the C184S mutant (Fig. 4D and E). However, since HRas is expressed under high expression CMV promoter and in the absence of other regulatory elements, the overexpressed constructs do show a dominant effect over the endogenous HRas, showing cysteine mutant dependent inhibition of differentiation of myoblasts after treatment with 15d-PGJ2 (Fig. 4D and E).

Moreover, in this specific experiment, it is difficult to interpret without a control with no HRas construct and another without the 15d-PGJ2 treatment.

The mRNA levels of MyoD, MyoG, and MHC in C2C12 cells expressing HRas constructs after treatment with 15d-PGJ2 were normalized to the mRNA levels in C2C12 cells expressing corresponding constructs and were treated with vehicle (DMSO). mRNA levels in C2C12 cells treated with vehicle were not shown as they were normalized to 1. MHC protein levels in C2C12 cells expressing HRas constructs after 15d-PGJ2 treatment were normalized to that in C2C12 cells treated with vehicle (DMSO). Since the hypothesis to study the effect of HRas cysteine mutations on the differentiation of myoblasts after treatment with 15d-PGJ2, C2C12 cells expressing HRas WT serve as adequate control. Fig. 2 shows the effect of 15d-PGJ2 on muscle differentiation when HRas was not overexpressed.

Moreover, the overall study does not delineate the toxic effects of 15d-PGJ2 from its effect on the differentiation.

The inhibition of differentiation in C212 cells after treatment with 15d-PGJ2 cannot be attributed to the general toxicity of 15d-PGJ2 in cells. We show that the inhibition of differentiation of myoblasts after 15d-PGJ2 depends on modification of HRas at C184 i.e. failure to modify HRas at C184 (Fig. 3A) and resultant activation (Fig. 3B) by 15d-PGJ2 rescues this inhibition of differentiation of C2C12 cells (Fig. 4D and E), dissecting the inhibition of differentiation of myoblasts by 15d-PGJ2 from general toxic effects of 15d-PGJ2 on cell physiology.

Please note that the effect of 15d-PGJ2 on cell physiology is context-specific. On one hand, 15d-PGJ2 has been shown to exert tumor-suppressor effects by inhibiting the proliferation of ovarian cancer cells and lung adenocarcinoma cells (de Jong et al., 2011; Slanovc et al., 2024), 15d-PGJ2 also exerts pro-carcinogenic effects by induction of epithelial to mesenchymal transition in breast cancer cells MCF7 and inhibition of tumor-suppressor protein p53 in MCF7 and PC-3 cells (Choi et al., 2020; Kim et al., 2010).

Reviewer #2 (Public Review):

Summary:

In this study, Swarang and colleagues identified the lipid metabolite 15d-PGJ2 as a potential component of senescent myoblasts. They proposed that 15d-PGJ2 inhibits myoblast proliferation and differentiation by binding and regulating HRas, suggesting its potential as a target for restoring muscle homeostasis post-chemotherapy.

Strengths:

The regulation of HRas by 15d-PGJ2 is well controlled.

Weaknesses:

(1) I still think the novelty is limited by previous published findings. The authors themselves noted that the accumulation of 15d-PGJ2 in senescent cells has been reported in various cell types, including human fibroblasts, HEPG2 hepatocellular carcinoma cells, and HUVEC endothelial cells (PMCID: PMC8501892). Although the current study observed similar activation of 15d-PGJ2 in myoblasts, it appears to be additive rather than fundamentally novel. The covalent adduct of 15d-PGJ2 with Cys-184 of H-Ras was reported over 20 years ago (PMID: 12684535), and the biochemical principles of this interaction are likely universal across different cell types. The regulation of myogenesis by both HRas and 15d-PGJ2 has also been previously extensively reported (PMID: 2654809, 1714463, 17412879, 20109525, 11477074). The main conceptual novelty may lie in the connection between these points in myoblasts. But as discussed in another comment, the use of C2C12 cells as a model for senescence study is questionable due to the lack of the key regulator p16. The findings in C2C12 cells may not accurately represent physiological-relevant myoblasts. It is recommended that these findings be validated in primary myoblasts to strengthen the study's conclusions.

This is the first study to show a molecular mechanism where activation of HRas signaling in skeletal myoblasts due to covalent modification by 15d-PGJ2 at C184 of HRas inhibits the differentiation of skeletal myoblasts.

(2) The C2C12 cell line is not an ideal model for senescence study.

C2C12 cells are a well-established model for studying myogenesis. However, their suitability as a model for senescence studies is questionable. C2C12 cells are immortalized and do not undergo normal senescence like primary cells as C2C12 cells are known to have a deleted p16/p19 locus, a crucial regulator of senescence (PMID: 20682446). The use of C2C12 cells in published studies does not inherently validate them as a suitable senescence model. These studies may have limitations, and the appropriateness of the C2C12 model depends on the specific research goals.

Several reports have shown that cells undergo senescence independent of p16 expression. MCF7 human breast adenocarcinoma cells have been shown to undergo DNA damage mediated and Oncogene induced senescence as seen after treatment with Doxorubicin (PMID: PMC7025418) and expression of constitutively active HRas (PMID: 17135242), despite the homozygous deletion of p16 locus (ISBN 9780124375512 Chapter 17 Table 2) by upregulation of cell cycle inhibitor protein p21. In this study, we observe an increase in the senescence markers in C2C12 cells after treatment with Doxo (Fig. 1). We also observed an increase in the markers of DNA damage-mediated senescence in MCF7 after treatment with Doxo (Data will be included in the revised manuscript). Based on these observations, we have concluded that C2C12 cells undergo senescence despite lacking the p16/p19 locus.

In the study by Moustogiannis et al. (PMID: 33918414), they claimed to have aged C2C12 cells through multiple population doublings. However, the SA-β-gal staining in their data, which is often used to confirm senescence, showed almost fully confluent "aged" C2C12 cells. This confluent state could artificially increase SA-β-gal positivity, suggesting that these cells may not truly represent senescence. Moreover, the "aged" C2C12 cells exhibited normal proliferation, which contradicts the definition of senescence. Similar findings were reported in another study of C2C12 cells subjected to 58 population doublings (PMID: 21826704), where even at this late stage, the cells were still dividing every 2 or 3 days, similar to younger cells at early passages. More importantly, I do know how the p16 was detected in that paper since the locus was already mutated. In terms of p21, there was no difference in the proliferative C2C12 cells at day 0.

In the study by Moiseeva et al. in 2023 (PMID: 36544018), C2C12 cells were used for senescence modeling for siRNA transfection. However, the most significant findings were obtained using primary satellite cells or confirmed with complementary data.

In conclusion, while molecular changes observed in studies using C2C12 cells may be valid, the use of primary myoblasts is highly recommended for senescence studies due to the limitations and questionable senescence characteristics of the C2C12 cell line.

(3) Regarding source of increased PGD in the conditioned medium, I want to emphasize that it's unclear whether the PGD or its metabolites increase in response to DNA damage or the senescence state. Thus, using a different senescent model to exclude the possibility of DNA damage-induced increase will be crucial.

Though Senescence can be induced by several stress stimuli like DNA damage, Oncogene expression, ROS, Mitochondrial Dysfunction, etc., DNA damage remains critical for the induction of the SASP (reviewed in PMID: 20078217). Also, other models of senescence, like Oncogene Induced Senescence (reviewed in PMID: 17671427), ROS Induced Senescence (PMID: 24934860), Mitochondrial Dysfunction Associated Senescence (MiDAS) (PMID: 26686024) have shown upregulation of DNA damage-associated signaling pathways. In this study, we have explored the SASP of cells undergoing senescence upon chemotherapy drug Doxorubicin-mediated DNA damage.

(4) Similarly for the in vivo Doxorubicin (Doxo) injection, both reviewers have raised concerns about the potential side effects of Doxo, including inflammation, DNA damage, and ROS generation. These effects could potentially confound the results of the study. The physiological significance of this study will heavily rely on the in vivo data. However, the in vivo senescence component is confounded by the side effects of Doxo.

We concur that this is a limitation of this study and the subsequent work will demonstrate the origin of prostaglandin biosynthesis after treatment with Doxo in vivo.

(5) Figure 2A lacks an important control from non-senescent cells during the measurement of C2C12 differentiation in the presence of conditioned medium. The author took it for granted that the conditioned medium from senescent cells would inhibit myogenesis, relying on previous publications (PMID: 37468473). However, that study was conducted in the context of myotonic dystrophy type 1. To support the inhibitory effect in the current experimental settings, direct evidence is required. It would be necessary to include another control with conditioned medium from normal, proliferative C2C12 cells.

Conditioned medium of senescent cells of several types, like senescent myoblasts in case of DM1 (PMID: 37468473), adipocytes undergoing senescence due to H2O2 treatment, Insulin Resistance, and Replicative senescence (PMID: 37321332), has been shown to inhibit the differentiation of myoblasts. Therefore, in this study, we measured the effect of prostaglandin PGD2 and its metabolites on the differentiation of myoblasts by inhibiting the biosynthesis of PGD2 in senescent myoblasts by treatment with AT-56. We inhibited the synthesis of PGD2 in senescent cells by treatment with AT-56, and then collected the conditioned medium. Conditioned medium collected from senescent C2C12 cells treated with vehicle (DMSO) served as a control for the experiment.

(6) Statistical analyses problems.

Only t-test was used throughout the study even when there are more than two groups. Please have a statistician to evaluate the replicates and statistical analyses used.

In experiments with more than two groups, the t-test was used for column-wise comparison of the experiment samples to the control sample. Multiple sample comparisons using one-way or two-way ANOVA were avoided as experimental samples were individually compared to the control sample.

For the 15d-PGJ2/cell concentration measurements in Figure 1F, there were only two replicates, which was provided in the supplementary table after required. Was that experiment repeated with more biological replicates?

Additional replicates of the experiment will be included in the revised manuscript.

For figure 1C, Fig 1F, 1G, 1J, 2C, 2E, 3A, 3E, 3F, 4D, 4E, please include each data points in bar graphs as used in Fig 1D, or at least provide how many biological replicates were used for each experiment?

Appropriate revisions will be made in the figure legends of the revised manuscript.

There is no error bar in a lot of control groups (Fig 2C, 2E, 3EF, 4E, S4B).

There are no error bars for the control groups in the figures 2C, 2E, 3E, 3F, 4E, and S4B as the experimental samples of each replicate were normalized to the corresponding control sample, rendering the values for the control sample of each replicate to 1.

For qPCR data in Figure 1C, the author responded in that the data in was plotted using 2-ΔCT instead of 2-ΔΔCT to show the variability in the expression of mRNAs isolated from animals treated with Saline. This statement does not align with the method section. Please revise.

Appropriate revisions will be made to the method sections of the revised manuscript.

(7) For Figure 1, the title may not be appropriate as there is insufficient data to support the inhibition of myoblast differentiation.

Appropriate revisions will be made to the revised manuscript.

Recommendations for the authors:

After careful review, the editors advise you to carefully address the following concerns.

(1) There were concerns that in the revised manuscript, the DMSO and Doxo experiments depicted in Figure 1H appeared quite homogenous despite the author's description to the contrary. This leads to concerns about the type of statistics employed and the possible low number of replicates of experiments shown in Fig. 1.

(2) Experiments in Figure 1F, 1I, and 1J had as few as n=2 experiments. Figures 1C, 1D, 1F, 1G, and 1J, the statistics used a two-tailed student's t-test; for all other experiments, they marked N/A for statistics. Using a t-test for multi-group comparisons (as indicated in the figure legend) and relying on only 2 replicates for many experiments are not appropriate.

Additional replicates for the experiments shown in figures 1F, 1I, and 1J have been done and the data will be revised along with updated statistical tests during the revision of the manuscript.

(3) In several experiments, the difference between technical replicates is too high.

Reviewer #1 (Recommendations For The Authors):

Most of my concerns were addressed in the revised manuscript.

We thank the reviewer for their time in reviewing the manuscript and consideration of the author’s response to their comments in during the previous round of review.

Reviewer #2 (Recommendations For The Authors):

Validating the findings in a primary myoblast is highly recommended for senescence studies due to the limitations and questionable senescence characteristics of the C2C12 cell line.

We have explained the statistical tests used in the manuscript in the general comment section of the reviewer’s comments.

Validate the finding in a different senescent model to exclude the possibility of DNA damage-response.

We have explained the statistical tests used in the manuscript in the general comment section of the reviewer’s comments.

For Fig 2A, add another control with a conditioned medium from normal, proliferative C2C12 cells.

We have explained the statistical tests used in the manuscript in the general comment section of the reviewer’s comments.

Please have a statistician to evaluate the replicates and statistical analyses used.

We have explained the statistical tests used in the manuscript in the general comment section of the reviewer’s comments.

For the barplots (figure 1C, Fig 1F, 1G, 1J, 2C, 2E, 3A, 3E, 3F, 4D, 4E), please include each data points, or at least provide how many biological replicates were used for each experiment.

Appropriate revisions will be made in the figure legends of the revised manuscript.

For Figure 1, the title may not be appropriate as there is insufficient data to support the inhibition of myoblast differentiation.

Appropriate revisions will be made to the revised manuscript.

The following is the authors’ response to the original reviews.

eLife assessment

This manuscript provides useful information about the lipid metabolite 15d-PGJ2 as a potential regulator of myoblast senescence. The authors provide experimental evidence that 15d-PGJ2 inhibits myoblast proliferation and differentiation by binding and regulating HRas. However, the manuscript is incomplete in its current form, as it lacks robust support from the data regarding the main conclusions related to senescence and technical concerns related to the senescence models used in this study.

We are grateful to the editors and the reviewers for their time and comments in sharpening the science and the writing of the manuscript. We have attached a detailed response to emphasize that the manuscript does include robust evidence regarding the claims, which could have been missed during the review process. We have provided a better context for these points now.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors show that upon treatment with Doxorubicin (Doxo), there is an increase in senescence and inflammatory markers in the muscles. They also show these genes get upregulated in C2C12 myoblasts when treated with conditioned media or 15d-PGJ2. 15dPGJ2 induces cell death in the myoblasts, decreases proliferation (measured by cell numbers), and decreases differentiation and fusion. 15d-PGJ2 modified Cys184 of HRas, which is required for its activation as indicated by the FRET analysis with RAF RBD. They also showed that 15d-PGJ2 activates ERK signaling, but not Akt signaling, through the electrophilic center. 15d-PGJ2 inhibits Golgi localization of HRAS (only WT, not C181 or C184 mutant). They also showed that expressing the WT HRas followed by 15d-PGJ2 treatment led to a decrease in the levels of MHC mRNA and protein, and this defect is dependent on C184. This is a well-written manuscript with interesting insights into the mechanism of action of 15d-PGJ2. However, some clarification and experiments will help the paper advance the field significantly.

Strengths:

The data clearly shows that 15d-PGJ2 has a negative role in the myoblast cells and that it leads to modification of HRas protein. Moreover, the induction of biosynthetic enzymes in the PGD2 pathway also supports the induction of 15d-PGJ2 in Doxorubicin-treated cells. Both conditioned media experiments and the 15d-PGJ2 experiments show that 15d-PGJ2 could be the active component secreted by the senescent myoblasts.

Weaknesses:

The genes that are upregulated in the muscles upon injection with Doxo are also markers for inflammation. Since Doxo is also known to induce systemic inflammation, it is important to delineate these two effects (inflammatory cells vs senescent cells). The expression of beta Gal and other markers of senescence in the tissue sections will help to delineate these.

As pointed out Doxo induces systemic inflammation along with inducing DNA damage-mediated senescence. Therefore, along with the inflammatory markers of the SASP (CXCL1/2, TNF1α, IL6, PTGS1/2, PTGDS) we also observed an increase in the mRNA levels of canonical markers of DNA damage-mediated senescence. We observed an increase in the mRNA levels of cell cycle and senescence associated proteins p16 and p21 (Fig. 1C). We also observed an increased nuclear accumulation of p21 (Fig. 1A) and increased levels of phosphorylated H2A.X in the nucleus (Fig. 1B).

In Figure 2, where the defect in the differentiation of myoblasts upon treatment with 15d-PGJ2 is shown, most of the cells die within 48 hours at higher concentrations, making it difficult to perform the experiments. This also shows that 15d-PGJ2 was toxic to these cells. Lower concentrations show a decrease in the differentiation based on the lower number of nuclei in fibers and low expression of MyoD, MyoG, and MHC. However, it is unclear if this is due to increased cell death or defective differentiation. It would be a lot more informative if the cell count, cell division, and cell death could be plotted for these concentrations of the drug during the experiment.

We measured the viability of C2C12 cells after 24 hours of treatment with 15d-PGJ2 using the MTT assay and observed that the viability of cells was decreased after treatment with 15d-PGJ2 (10 µM) but not with 15d-PGJ2 (1 µM, 2 µM, 4 µM, or 5 µM) (see Fig. S2A of the updated manuscript). The results and figures of the manuscript have been updated accordingly.

Also, in the myoblast experiments, are the effects of treatment with Dox reversible?

The treatment with Doxorubicin is irreversible as the senescent phenotype was not reversed after withdrawal of Doxorubicin, even after 20 days.

In Figure 3, most of the experiments are done at a high concentration, which induces almost complete cell death within 48 hours.

Figure 3 is an acute experiment for only 1 hour, at which time no cell death was observed. Specifically, we measured the phosphorylation of Erk and Akt proteins after 1 hour of treatment with 15d-PGJ2 (10 µM) during which we did not observe any cell death.

Even at such a high concentration of 15dPGJ2, the increase in ERK phosphorylation is minimal.

We observe a ~30% increase in the phosphorylation of Erk proteins after treatment with 15d-PGJ2 in 0.2% serum medium compared to treatment with vehicle (DMSO). This is reproducible and significant.

The experiment Figure 4C shows that C181 and C84 mutants of the HRas show higher levels in Golgi compared with WT. However, this could very well be due to the defect in palmitoylation rather than the modification with 15d-PGJ2.

Our data does not suggest higher levels of C184S mutant in the Golgi compared with WT (Fig. S4A). We observed that the ratio of HRas levels in the Golgi to the HRas levels in the plasma membrane were similar in C2C12 cells expressing HRas C184S and HRas WT (Fig. S4A graph columns 1 and 5).

Though the authors allude to the possibility that intracellular redistribution of HRas by 15d-PGJ2 requires C181 palmitoylation, the direct influence of C184 modification on C181 palmitoylation is not shown. To have a meaningful conclusion, the authors need to compare the palmitoylation and modification with 15d-PGJ2.

Palmitoylation of HRas C181S is required for the localization of HRas at the plasma membrane. The inhibition of palmitoylation of C181, either by mutation (C181S) or treatment with protein palmitoyl transferase inhibitor (2-Bromopalmitate), results in the accumulation of HRas at Golgi(Rocks et al., 2005) (Fig. S4A). Modification of HRas at C184 by 15d-PGJ2 (Fig. 3A) could inhibit the palmitoylation of HRas at C181. However, our data does not support this hypothesis as modification of HRas WT by 15d-PGJ2 does not increase the level of HRas at the Golgi, like in the case of inhibition of cysteine palmitoylation due to C181S mutation.

To test if the inhibition of myoblast differentiation depends on HRas, they overexpressed the HRas and mutants in the C2C12 lines. However, this experiment does not take the endogenous HRAs into consideration, especially when interpreting the C184 mutant. An appropriate experiment to test this would be to knock down or knock out HRas (or make knock-in mutations of C184) and show that the effect of 15d-PGJ2 disappears.

Endogenous HRas (wild type) is present in the C2C12 cells overexpressing the EGFP-tagged HRas constructs. Therefore, we only observe a partial rescue in the differentiation after 15d-PGJ2 treatment in C2C12 cells expressing the C184S mutant (Fig. 4D and E). However, since HRas is expressed under high expression CMV promoter and in the absence of other regulatory elements, the overexpressed constructs do show a dominant effect over the endogenous HRas, showing cysteine mutant dependent inhibition of differentiation of myoblasts after treatment with 15dPGJ2 (Fig. 4D and E).

Moreover, in this specific experiment, it is difficult to interpret without a control with no HRas construct and another without the 15d-PGJ2 treatment.

The mRNA levels of MyoD, MyoG, and MHC in C2C12 cells expressing HRas constructs after treatment with 15d-PGJ2 were normalized to the mRNA levels in C2C12 cells expressing corresponding constructs and were treated with vehicle (DMSO). mRNA levels in C2C12 cells treated with vehicle were not shown as they were normalized to 1. MHC protein levels in C2C12 cells expressing HRas constructs after 15d-PGJ2 treatment were normalized to that in C2C12 cells treated with vehicle (DMSO). Since the hypothesis to study the effect of HRas cysteine mutations on the differentiation of myoblasts after treatment with 15d-PGJ2, C2C12 cells expressing HRas WT serve as adequate control. Fig. 2 shows the effect of 15dPGJ2 on muscle differentiation when HRas was not overexpressed.

Moreover, the overall study does not delineate the toxic effects of 15d-PGJ2 from its effect on the differentiation.

The inhibition of differentiation in C212 cells after treatment with 15d-PGJ2 cannot be attributed to the general toxicity of 15d-PGJ2 in cells. We show that the inhibition of differentiation of myoblasts after 15d-PGJ2 depends on modification of HRas at C184 i.e. failure to modify HRas at C184 (Fig. 3A) and resultant activation (Fig. 3B) by 15d-PGJ2 rescues this inhibition of differentiation of C2C12 cells (Fig. 4D and E), dissecting the inhibition of differentiation of myoblasts by 15d-PGJ2 from general toxic effects of 15d-PGJ2 on cell physiology.

Please note that the effect of 15d-PGJ2 on cell physiology is context-specific. On one hand, 15d-PGJ2 has been shown to exert tumor-suppressor effects by inhibiting the proliferation of ovarian cancer cells and lung adenocarcinoma cells (de Jong et al., 2011; Slanovc et al., 2024), 15d-PGJ2 also exerts pro-carcinogenic effects by induction of epithelial to mesenchymal transition in breast cancer cells MCF7 and inhibition of tumor-suppressor protein p53 in MCF7 and PC-3 cells (Choi et al., 2020; Kim et al., 2010).

Reviewer #2 (Public Review):

Summary:

In this study, Swarang and colleagues identified the lipid metabolite 15d-PGJ2 as a potential component of senescent myoblasts. They proposed that 15d-PGJ2 inhibits myoblast proliferation and differentiation by binding and regulating HRas, suggesting its potential as a target for restoring muscle homeostasis post-chemotherapy.

Strengths:

The regulation of HRas by 15d-PGJ2 is well controlled.

Weaknesses:

The novelty of the study is compromised as the activation of PGD and 15d-PGJ2, as well as the regulation of HRas and cell proliferation, have been previously reported.

Literature does not support this statement, and it is important to clarify this misimpression for the field as a whole.

Let us clarify-

Covalent modification of HRas by 15d-PGJ2 has been reported only twice in the literature(Luis Oliva et al., 2003; Yamamoto et al., 2011) in fibroblasts and neurons respectively.

Interaction between Hras and 15d-PGJ2 in skeletal muscles has not been shown before, even though both Hras and 15d-PGJ2 are shown to be key regulators of muscle homeostasis.

Activation of Hras by 15d-PGJ2 was reported first by Luis Oliva et al (Luis Oliva et al., 2003). However, this study does not comment on the functional implications of activation of Hras signaling.

Recently, our lab contributed to a study where the functional implication of activation of Hras signaling due to covalent modification by 15d-PGJ2 was shown in the maintenance of senescence phenotype (Wiley et al., 2021).

15d-PGJ2 was shown to inhibit the differentiation of myoblasts by Hunter et al (Hunter et al., 2001). This study hypothesized that the inhibition of myoblast differentiation is via 15d-PGJ2 mediated activation of the PPARγ signaling, the study also showed inhibition of myoblast differentiation independent of PPARγ activity, suggesting the presence of other mechanisms.

This is the first study to show a molecular mechanism where activation of Hras signaling in skeletal myoblasts due to covalent modification by 15d-PGJ2 at C184 of Hras inhibits the differentiation of skeletal myoblasts.

Additionally, there are major technical concerns related to the senescence models, limiting data interpretation regarding the relevance to senescent cells.

Major concerns:

(1) The C2C12 cell line is not an ideal model for senescence study due to its immortalized nature and lack of normal p16 expression. A more suitable myoblasts model is recommended, with a more comprehensive characterization of senescence features.

C2C12 is a good model for DNA damage-based senescence that is used in this manuscript. Several reports in the literature have shown the induction of senescence in C2C12 cells. Moiseeva et al 2023 show induction of senescence in C2C12 cells after etoposide-mediated DNA damage. Moustogiannis et al 2021 show the induction of replicative senescence in C2C12 cells. In this study, we show that C2C12 cells undergo DNA damage-mediated senescence after treatment with Doxo. We measured the induction of senescence in C2C12 cells upon DNA damage using several physiological (Nuclear Size, Cell Size, and SA β-gal) and molecular markers (mRNA levels of p21 and SASP factors (IL6 and TGFβ), protein levels of p21) of senescence (see Fig. 1 of the updated manuscript). The results and the figures in the manuscript have been updated accordingly.

(2) The source of increased PGD or its metabolites in the conditioned medium is unclear. Including other senescence models, such as replicative or oncogeneinduced senescence, would strengthen the study.

Fig. 1E shows time-dependent increase in the expression of PGD2 biosynthetic enzymes in senescent C2C12 cells. Fig. 1F shows an increase in the levels of 15dPGJ2 secreted by senescent C2C12 cells in the conditioned medium. This data shows that senescent C2C12 cells are the source of PGD and its metabolites in the conditioned medium.

Again, C2C12 is not suitable for replicative senescence due to its immortalized status.

We and others have shown that C2C12 cells undergo senescence, and this manuscript only used DNA damage induced senescence.

(3) In the in vivo part, it is unclear whether the increased expression of PTGS1, PTGS2, and PTGDS is due to senescence or other side effects of DOXO.

We concur that this is a limitation of this study and the subsequent work will demonstrate the origin of prostaglandin biosynthesis after treatment with Doxo in vivo.

(4) Figure 2A lacks an important control from non-senescent cells during the measurement of C2C12 differentiation in the presence of a conditioned medium.

Figure 2A tests the effect of prostaglandin PGD2 and its metabolites secreted by the senescent cells on the differentiation of myoblasts. Therefore, we inhibited the synthesis of PGD2 in senescent cells by treatment with AT-56, and then collected the conditioned medium. Conditioned medium collected from senescent C2C12 cells treated with vehicle (DMSO) served as a control for the experiment, whereas differentiation of C2C12 cells without any treatment serves as a positive control.

There is no explanation of how differentiation was quantified or how the fusion index was calculated.

The fusion index was calculated using a published myotube analyzer software (Noë et al., 2022). Appropriate information has been added to the materials and methods section of the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

Line 3: Expand SA in "SA β-gal".

The manuscript has been updated accordingly (See line 3).

Line 68: HRas is highly regulated by lipid modifications.

The manuscript has been updated accordingly (See line 67).

Figures

Figure S1A seemed incomplete (maybe some processing issue).

The Figure has been updated in the revised manuscript (See Fig. S1A).

Figure S1B-H are mislabeled.

The figure has been updated in the revised manuscript (See Fig. S1C, D, E, and F).

Figures S1E-H are not mentioned in the manuscript.

The manuscript has been updated accordingly (See line 120).

Many supplementary figures are not cited in the article.

The manuscript has been updated accordingly. (See lines 85, 120, 123, 166, 225, 356, 364, 412, and 413)

Reviewer #2 (Recommendations For The Authors):

(1) Clarify the injection method for Doxorubicin in B6J mice on line 83 (IP or IM).

Mice were injected intraperitoneally with Doxorubicin (as mentioned in the materials and methods, see lines 83 and 794)

(2) Address missing information in figures or figure legends.

There is missing piece in Sup Fig 1A.

The figure has been updated in the revised manuscript (See Fig. S1A).

Correct labels in Sup Fig 1C and 1D.

The figure has been updated in the revised manuscript (See Fig. S1C, D, E, and F).

How would the authors explain the dramatic differences in the morphology of C2C12 cells treated with DOXO between bright field and SA-beta-gal staining images in Sup Fig 1B and 1C.

The SA β-gal image after treatment with Doxo does show a flattened cell morphology. Another field of view from the same experiment has been added in the figure to show the difference in the cell morphology more prominently in the revised manuscript (See Fig. 1H).

Provide explanations for Sup Fig 1E-1G, including the meaning of the y-axis and the blue dots and red lines.

We have provided an explanation for the multiple reaction monitoring mass spectrometry used to measure the concentration of 15d-PGJ2 in the conditioned medium in the revised manuscript (see lines 119-130 and the legends of Fig. S1C, D, and E)

(3) Please review the calculation of qPCR data in Figure 1C for correctness, ensuring reference samples with an average expression level of 1.

The data in Fig. 1C was plotted using 2-ΔCT instead of 2-ΔΔCT to show the variability in the expression of mRNAs isolated from animals treated with Saline.

(4) Please explain the calculation of 15d-PGJ2/cell concentration in Figure 1F and provide raw data for review, considering the substantial changes and small error bars. The method or result section lacks an explanation of how this calculation was performed. Additionally, there is no mention of the cell number count.

All the raw values (concentration of 15d-PGJ2 measured using mass spec and cell numbers counted at the time of collection of conditioned medium) are provided in the supplementary table 1. The standard curve to calculate the concentration of 15dPGJ2 in the conditioned medium is shown in Fig. S1F. The cell number was counted after trypsinization using a hemocytometer on the day of collection of the conditioned medium.

(5) Please clarify how cell number normalization and doubling time calculation were done in Fig 2B. Consider replacing the figure with a growth curve showing confluence on the y-axis for easier interpretation.

Cells were counted every 24 hours and the normalization was done to the number of cells counted on day 0 of the treatment (to consider attaching efficiency and other cell culture parameters). Doubling time was calculated as the reciprocal of the slope of the graph of log2(normalized cell number) vs time.

Version 2

Abstract

Abstract

Senescent cells, which are characterized by multiple features such as increased expression of Senescence-Associated β-galactosidase activity (SA β-gal) and cell cycle inhibitors such as p21 or p16, accumulate with tissue damage and dysregulate tissue homeostasis. In the context of skeletal muscle, it is known that agents used for chemotherapy such as Doxorubicin cause buildup of senescent cells, leading to the inhibition of tissue regeneration. Senescent cells influence the neighboring cells via numerous secreted factors which form the senescence-associated secreted phenotype (SASP). Lipids are emerging as a key component of SASP that can control tissue homeostasis. Arachidonic acid-derived lipids have been shown to accumulate within senescent cells, specifically 15d-PGJ2, which is an electrophilic lipid produced by the non-enzymatic dehydration of the prostaglandin PGD2. In this study, we show that 15d-PGJ2 is also released by Doxorubicin-induced senescent cells as a SASP factor. Treatment of skeletal muscle myoblasts with the conditioned medium from these senescent cells inhibits myoblast fusion during differentiation. Inhibition of L-PTGDS, the enzyme that synthesizes PGD2, diminishes the release of 15d-PGJ2 by senescent cells and restores muscle differentiation. We further show that this lipid post-translationally modifies Cys184 of HRas in skeletal muscle cells, causing a reduction in the localization of HRas to the Golgi, increased HRas binding to RAF RBD, and activation of cellular MAPK-Erk signaling (but not the Akt signaling). Mutating C184 of HRas prevents the ability of 15d- PGJ2 to inhibit the differentiation of muscle cells and control the activity of HRas. This work shows that 15d-PGJ2 released from senescent cells could be targeted to restore muscle homeostasis after chemotherapy.

Introduction

Senescent cells are important drivers of aging and damage-associated loss of tissue homeostasis(Childs et al., 2015). Anti-cancer chemotherapy presents an important context where treatment with chemotherapeutics such as Doxorubicin (Doxo) causes widespread cellular senescence which inhibits tissue homeostasis and regeneration, including in skeletal muscles(Francis et al., 2022). It has been shown that Doxo causes systemic inflammation and leads to the emergence of senescent cells across tissues(Di Leonardo et al., 1994; Hu and Zhang, 2019; Robles and Adami, 1998). Senescent cells negatively affect tissue homeostasis and regeneration by releasing factors including proteins like growth factors, matrix metalloproteases, cytokines, and chemokines, and small molecules like fatty acid derivatives(Campisi, 2005; Coppé et al., 2010; Dilley et al., 2003; Krtolica et al., 2001; Parrinello et al., 2005; Shelton et al., 1999; Yang et al., 2006). The release of these factors from senescent cells is called the Senescence-Associated Secretory Phenotype (SASP). It is expected that these SASP factors and their mechanisms of action will vary depending on cellular and tissue contexts. Identifying SASP factors and their underlying mechanistic targets will be critical for building an understanding of how senescent cells control tissue homeostasis(Coppé et al., 2010; Davalos et al., 2010). Lipids are a less explored family of SASP factors, and it is important to understand how they affect tissue regeneration(Hamsanathan and Gurkar, 2022). We have previously shown that senescent cells have increased intracellular levels of prostaglandin 15d-PGJ2(Wiley et al., 2021), a non-enzymatic dehydration product of prostaglandin PGD2(Shibata et al., 2002). In the context of skeletal muscle, PGD2 and 15d-PGJ2 have been shown to negatively regulate muscle differentiation via mechanisms that do not depend on a cognate receptor(Hunter et al., 2001; Veliça et al., 2010). Here we study the role of 15d-PGJ2 as a member of the SASP and identify the mechanisms by which it might negatively affect muscle regeneration. 15d-PGJ2 has been previously shown to covalently modify multiple proteins like MAPK1, MCM4, EIF4A-I, PKM1, GFAP etc. in endothelial and neuronal cells(Marcone and Fitzgerald, 2013; Yamamoto et al., 2011). 15d-PGJ2 was shown to be covalently modifying HRas in NIH3T3, Cos1, and IMR90 cell lines(Luis Oliva et al., 2003; Wiley et al., 2021). We further studied HRas as an important target that might mediate the effects of 15d-PGJ2 on muscle differentiation via covalent modification. We investigated HRas as a possible effector of 15d-PGJ2 because (i) HRas belongs to the Ras superfamily of small molecule GTPases and is a known regulator of key cellular processes(Davis et al., 1983; Harvey, 1964; Kirsten and Mayer, 1967; Vetter and Wittinghofer, 2001). (ii) constitutively active HRas mutant (HRas V12) has been shown to inhibit the differentiation of myoblasts by inhibiting MyoD and Myogenin expression(Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Van Der Burgt et al., 2007). (iii) Downstream signaling of HRas is important for muscle homeostasis as skeletal and cardiac myopathies are observed in individuals carrying constitutively active mutants of HRas(Engler et al., 2021; Konieczny et al., 1989; Lee et al., 2010; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007). HRas is highly regulated by lipid modifications, it undergoes reversible palmitoylation and de-palmitoylation at C-terminal cysteines, which regulate the intracellular distribution and activity of HRas(Gutierrez et al., 1989; Lu and Hofmann, 1995; Rocks et al., 2005). In this study, we show that 15d-PGJ2 is synthesized and released by senescent myoblasts upon treatment with Doxo. 15d-PGJ2, taken up by the myoblasts, covalently modifies HRas at cysteine 184 and activates it. We also show that previously reported inhibition of differentiation of myoblasts by 15d-PGJ2 depends on HRas C-terminal cysteines, notably cysteine 184. This study provides a mechanism by which prostaglandins secreted as SASP inhibit the differentiation of myoblasts, affecting muscle homeostasis in patients undergoing chemotherapy.

Results

Doxorubicin (Doxo) treatment induces senescence in mouse skeletal muscles and C2C12 mouse myoblasts

Doxorubicin-mediated DNA damage has been shown to induce senescence in cells(Di Leonardo et al., 1994; Hu and Zhang, 2019; Robles and Adami, 1998). Therefore, we injected B6J mice intraperitoneally with Doxorubicin (Doxo) (5mg/kg) every 3 days for 9 days and observed induction of DNA damage-mediated senescence in hindlimb skeletal muscles (Fig. S1A). We observed an increase in the expression of p21 and increased nuclear levels of the DNA damage marker γH2A.X in mouse Gastrocnemius muscles (Fig. 1A and B). We also observed a significant increase in the mRNA levels of known senescence markers (p16 and p21), SASP factors (CXCL1, CXCL2, TNF1α, IL6, TGFβ1) in skeletal muscles of mice treated with Doxo compared to that of mice treated with saline (Fig. 1C). These observations suggest that there is induction of senescence in skeletal muscles of mice upon treatment with Doxo.

Prostaglandin synthesis and release by Doxo-induced senescent cells inhibits myoblast differentiation.

A. Expression and localization of tumor suppressor protein p21, measured by immunofluorescence, in hindlimb skeletal muscles of mice after 11 days of treatment with Doxo (5 mg/kg) or Saline.

B. Representative confocal micrograph of expression of γH2A.X in the gastrocnemius muscle of mice treated with Doxo (5 mg/kg) or Saline.

C. Expression of mRNAs of senescence markers (p16 and p21), SASP factors (CXCL1, CXCL2, TNFα1, IL6, TGFβ1), and enzymes involved in the biosynthesis of prostaglandin PGD2/15d-PGJ2 (PTGS1, PTGS2, PTGDS), measured by qPCR, in hindlimb skeletal muscles of mice after 11 days of treatment with Doxo (5 mg/kg) or Saline.

D. A representative confocal micrograph and a scatter plot of the nuclear area of C2C12 myoblasts, measured by immunofluorescence, after 16 days of treatment with Doxo (150 nM) or DMSO.

E. A representative widefield micrograph of cell morphology in C2C12 myoblasts after 16 days of treatment with Doxo (150 nM) or DMSO.

F. Expression of mRNA of cell cycle inhibitor p21 and SASP factors (IL6 and TGFβ), measured by qPCR, in C2C12 myoblasts after 16 days of treatment with Doxo (150 nM) or DMSO.

G. Expression of cell cycle inhibitor p21, measured by immunoblot, in C2C12 myoblasts after 16 days of treatment with Doxo (150 nM) or DMSO.

H. Activity of Senescence Associated β-galactosidase (SA β-gal), measured by X-gal staining at pH∼6, in C2C12 myoblasts after 16 days of treatment with Doxo (150 nM) or DMSO.

I. Expression of mRNAs of prostaglandin biosynthetic enzymes, measured by qPCR, in C2C12 myoblasts after treatment with Doxo (150 nM) or DMSO.

J. Concentration of 15d-PGJ2 released from quiescent or senescent C2C12 cells.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

C2C12 cells have been shown to undergo senescence after DNA damage, as assessed by an increase in the levels of SA β-gal and known markers of SASP (IL1α, IL6, CCL2, CXCL2, CXCL10)(Moiseeva et al., 2022). We treated C2C12 myoblasts with Doxo (150 nM) and observed a significant increase in the size of the nuclei (Fig. 1D), flattened cell morphology with an increase in the cell size (Fig. 1E), a significant increase in the mRNA levels of cell cycle inhibitor p21 and SASP factors IL6 and TGFβ (Fig. 1F), a significant increase in the protein levels of p21 (Fig. 1G), and an increase in the levels of SA β-gal (Fig. 1H) in C2C12 cells treated with Doxo. These observations suggest that C2C12 cells undergo senescence upon treatment with Doxo.

Doxo-mediated senescence induces synthesis and release of 15d-PGJ2 in C2C12 myoblasts and mouse skeletal muscle

Synthesis of prostaglandins by senescent cells has previously been reported(Wiley et al., 2021; Wiley and Campisi, 2021). Specifically, levels of PGD2 and its metabolite 15d-PGJ2 have been shown to be significantly increased in senescent cells. Therefore, we measured the levels of mRNA of enzymes involved in the synthesis of PGD2/15d-PGJ2 (PTGS1, PTGS2, and PTGDS), in the gastrocnemius muscle of mice after treatment with Doxo. We observed a significant increase in the mRNA levels of PTGS1, PTGS2, and PTGDS enzymes in the skeletal muscle of mice treated with Doxo (Fig. 1C). We also observed a time-dependent increase in the mRNA levels of PTGS1, PTGS2, PTGDS, and PTGES enzymes in C2C12 cells treated with Doxo compared to Day 0 (Fig. 1I). Expression of enzyme PTGES was elevated on Day 4, whereas the expression of Prostaglandin D synthase (PTGDS) increased only after Day 8, reaching maximum expression on Day 12. These observations suggest an increase in the synthesis of prostaglandins in senescent cells.

15d-PGJ2 is a non-enzymatic dehydration product of PGD2(Shibata et al., 2002). We observed an increase in the mRNA levels of synthetic enzymes of 15d-PGJ2 in senescent C2C12 cells. Therefore, we measured the levels of 15d-PGJ2 released by senescent C2C12 cells using targeted mass spectrometry (Fig. S1C, D, E, and F). The concentration of 15d-PGJ2 was quantified by monitoring the transition of the m/z of ions from 315.100 → 271.100 using a SCIEX 6500 mass spectrometer. We plotted a standard curve using purified 15d-PGJ2 (Fig. S1F) to quantify the concentration of 15d-PGJ2. We used the representative peaks from the conditioned medium collected from C2C12 cells incubated in 0.2% serum medium for 3 days (Quiescent cells) and C2C12 cells treated with Doxo (150 nM) (Senescent cells) to measure the concentration of 15d-PGJ2 released by quiescent cells or senescent C2C12 cells. We observed a significant increase (∼100 fold) in the concentration of 15d-PGJ2 in the conditioned medium from senescent cells as compared to that in quiescent cells (Fig. 1J). This suggests that senescent C2C12 cells release 15d-PGJ2 in the medium.

Prostaglandin PGD2 and its metabolites in the conditioned medium of senescent cells inhibit the differentiation of C2C12 myoblasts

15d-PGJ2 (the final non-enzymatic dehydration product of PGD2) has been shown to inhibit the differentiation of myoblasts(Hunter et al., 2001). We observed the release of 15d-PGJ2 by senescent cells, showing that 15d-PGJ2 is a SASP factor (Fig. 1F). Conditioned medium of senescent cells inhibits the differentiation of myoblasts in myotonic dystrophy type 1(Conte et al., 2023). Therefore, we tested whether 15d-PGJ2, the terminal dehydration product of PGD2, is required for the inhibitory effect of SASP on the differentiation of myoblasts. We treated C2C12 myoblasts with the conditioned medium of senescent cells or senescent cells treated with 30 µM of AT-56 (a well-characterized inhibitor of prostaglandin D synthase (PTGDS))(Hu et al., 2021; S. Hu et al., 2023; Shunfeng Hu et al., 2023; Irikura et al., 2009) and measured the differentiation of myoblasts by calculating the fusion index. We observed a significant decrease (∼20%) in the fusion index of the C2C12 myoblasts treated with the conditioned medium of senescent cells (Fig. 2A), suggesting that SASP factors decrease the differentiation of myoblasts. This decrease in the inhibition was rescued in myoblasts treated with the conditioned medium of senescent cells treated with AT-56 (Fig. 2A). This suggests that prostaglandins PGD2/15d-PGJ2 released by senescent cells as SASP factors can inhibit the differentiation of myoblasts.

15d-PGJ2 inhibits differentiation of myoblasts.

A. Expression of MHC protein and the fusion of myoblasts in myotubes, measured by immunofluorescence, after treatment with conditioned medium of senescent cells treated with PTGDS inhibitor AT-56 (30 µM) or DMSO

B. Normalized number of C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO

C. Expression of mRNAs of markers of differentiation (MyoD, MyoG, and MHC), measured by qPCR, in C2C12 myoblasts treated with 15d-PGJ2 (1 µM, 2 µM, or 4 µM) or DMSO.

D. Expression of MHC protein and the fusion of myoblasts in syncytial myotubes, measured by immunofluorescence, after treatment with 15d-PGJ2 (4 µM) or DMSO.

E. Expression of MHC protein, measured by immunoblotting, in primary human skeletal myoblasts after treatment with 15d-PGJ2 (1 µM, 2 µM, or 4 µM) or DMSO for 5 days.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 inhibits the proliferation and differentiation of mouse and human myoblasts

15d-PGJ2 has been shown to affect the proliferation of cancer cell lines, both positively and negatively (Chen et al., 2003; Choi et al., 2020; Slanovc et al., 2024; Yen et al., 2014). We measured the effect of 15d-PGJ2 on the proliferation of C2C12 myoblasts. We treated C2C12 myoblasts with 15d-PGJ2 (10 µM) or DMSO in DMEM 10% Serum medium for 72 hours and observed a significant decrease in the proliferation of C2C12 cells after treatment with 15d-PGJ2 (Fig. 2B). The doubling time of C2C12 cells was also increased upon treatment with 15d-PGJ2 (57.24 hours) compared to DMSO (13.76 hours). This suggests that 15d-PGJ2 decreases the proliferation of C2C12 myoblasts.

We measured the differentiation of C2C12 mouse and primary human myoblasts after treatment with 15d-PGJ2. To rule out the toxic effects of 15d-PGJ2 on cell physiology, we treated C2C12 cells with 15d-PGJ2 (1 µM, 2 µM, 4 µM, 5 µM, and 10 µM) in the C2C12 differentiation medium, and measured the viability of cells after 24 hours of treatment, using an MTT viability assay. We judged that 15d-PGJ2 was not cytotoxic up to 5 µM in the C2C12 differentiation medium (Fig. S2A). Based on this, we treated differentiating myoblasts with 15d-PGJ2 (1 µM, 2 µM, and 4 µM) for 5 days to measure the effects of 15d-PGJ2 treatment on differentiation of myoblasts. We observed a dose-dependent decrease in the mRNA levels of MyoD, MyoG, and MHC in differentiating C2C12 cells after treatment with 15dPGJ2 (Fig. 2C). There was a significant decrease in the no. of nuclei in individual MHC+ve fiber (∼75%) in C2C12 cells treated with 15d-PGJ2 (4 µM) compared to DMSO (Fig. 2D), suggesting a decrease in the fusion of myoblasts in myotubes. We also observed a dose-dependent decrease in the protein levels of MHC in differentiating primary human myoblasts upon treatment with 15d-PGJ2 (Fig. 2E). Together, these observations suggest that 15d-PGJ2 inhibits the differentiation of both mouse and human myoblasts.

Biotinylated 15d-PGJ2 covalently modifies HRas at Cysteine 184

15d-PGJ2 has been shown to covalently modify several proteins including p53 and NF-κB, which are involved in several key biological processes(Marcone and Fitzgerald, 2013). HRas was identified to be covalently modified by 15d-PGJ2 at cysteine 184 in NIH3T3 and Cos1 cells(Luis Oliva et al., 2003). Therefore, we tested whether 15d-PGJ2 could covalently modify HRas in C2C12 cells. We treated C2C12 cells expressing the EGFP-tagged wild-type HRas with biotinylated 15d-PGJ2 (5 µM). We then immunoprecipitated biotinylated 15d-PGJ2 using streptavidin. We observed a significant increase (∼3.5 fold) in the pulldown of HRas upon treatment with 15d-PGJ2biotin compared to DMSO (Fig. 3A), suggesting an interaction between 15d-PGJ2 and HRas. To measure the role of individual C-terminal cysteines in the binding of HRas with 15d-PGJ2, we treated C2C12 cells expressing the EGFP-tagged C181S and C184S mutants of HRas with biotinylated 15d-PGJ2 (5 µM), and immunoprecipitated using streptavidin. We observed that the intensity of EGFP-tagged HRas was significantly decreased in cells expressing the C184S mutant (∼80% decrease compared to HRas WT) but not in those expressing the C181S mutant (Fig. 3A). This suggests that 15d-PGJ2 covalently modifies HRas at cysteine 184 in C2C12 cells.

15d-PGJ2 covalently modifies HRas at Cysteine 184 and activates the HRas-MAPK pathway via the electrophilic cyclopentenone ring.

A. Streptavidin-immunoprecipitation of EGFP-HRas, measured by immunoblotting, in C2C12 cells after 3 hours of treatment with 15d-PGJ2-Biotin (5 µM).

B. Representative confocal micrograph of Fluorescence Resonance Energy Transfer (FRET) between EGFP-tagged HRas (EGFP-HRas) and mCherry-tagged Ras binding domain (RBD) of RAF kinase (mCherry-RAF RBD).

C. Activation of the EGFP-tagged wild type HRas (HRas WT), measured by FRET, before and after 1 hour of treatment with 15d-PGJ2 (10 µM) after starvation for 24 hours.

D. Activation of the EGFP-tagged C-terminal cysteine mutants of HRas (HRas C181S and HRas C184S), measured by FRET, before and after 1 hour of treatment with 15d-PGJ2 (10 µM) after starvation for 24 hours.

E. Phosphorylation of Erk (42 kDa and 44 kDa), measured by immunoblotting, in C2C12 cells after 1 hour of treatment with 15d-PGJ2 (5 µM, 10 µM) or DMSO after starvation for 24 hours.

F. Phosphorylation of Erk (42 kDa and 44 kDa), measured by immunoblotting, in C2C12 cells after 1 hour of treatment with 15d-PGJ2 (10 µM)/ 9,10-dihydro-15d-PGJ2 (10 µM) or DMSO after starvation for 24 hours.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 increases the FRET between EGFP-HRas and mCherry-RAF-RBD in wild-type and C181S mutant but not in the C184S mutant of HRas

We next tested the effect of covalent modification of HRas by 15d-PGJ2 on HRas GTPase activity using FRET. mCherry-RAF-RBD is a well-characterized sensor of the activity of HRas. RAF-RBD binds to the activated HRas upon activation of HRas, allowing FRET between EGFP and mCherry(Rocks et al., 2005). We co-expressed EGFP-tagged HRas (EGFP-HRas) with mCherry-RAF-RBD in C2C12 myoblasts (Fig. 3B). We measured the efficiency of FRET between EGFP and mCherry using an ImageJ plugin, FRET analyzer(Hachet-Haas et al., 2006). We compared the mean acceptor normalized FRET index in C2C12 myoblasts co-expressing EGFP-HRas WT and mCherry-RAF-RBD before and after treatment of 15d-PGJ2 (10 µM) for 1 hour. We observed a significant increase (∼30%) in the mean acceptor normalized FRET index upon treatment with 15d-PGJ2 (Fig. 3C). This suggests that 15d-PGJ2 activates HRas. To measure the role of individual C-terminal cysteines in 15d-PGJ2 mediated activation of HRas, we co-expressed EGFP-HRas C181S or C184S with mCherry-RAF-RBD in C2C12 myoblasts. We measured the mean acceptor normalized FRET index before and after 1 hour of treatment with 15d-PGJ2 (10 µM). We observed a significant increase (∼40%) in the mean acceptor normalized FRET index in cells expressing EGFP-HRas C181S upon treatment with 15d-PGJ2 but not in cells expressing EGFP-HRas C184S (Fig. 3D). These observations suggest that activation of HRas by 15d-PGJ2 occurs in a cysteine 184 dependent manner.

15d-PGJ2 increases phosphorylation of Erk (Thr202/Tyr204) but not Akt (S473) in C2C12 myoblasts

HRas regulates two major downstream signaling pathways, the MAP kinase (MAPK) pathway and the PI3 kinase (PI3K) pathway(Pylayeva-Gupta et al., 2011). We tested the effects of treatment with 15d-PGJ2 on these two downstream signaling pathways by measuring the phosphorylation of Erk (42 kDa and 44 kDa) and Akt proteins in C2C12 cells. We treated C2C12 cells with 15d-PGJ2 (5 µM and 10 µM) or DMSO for 1 hr (after 24 hrs. of serum starvation) and observed a dose-dependent increase in the phosphorylation of Erk (T202/Y204) (42 kDa) but not of Erk (44 kDa) (Fig. 3E). We did not observe an increase in the phosphorylation of Akt (S473) in C2C12 cells after treatment with 15d-PGJ2 (Fig. S3C). These observations suggest that 15d-PGJ2 activates the MAPK signaling pathway, but not the PI3K signaling pathway.

15d-PGJ2 contains a reactive electrophilic center in its cyclopentenone ring, that can react with cysteine residues of proteins(Luis Oliva et al., 2003). We tested its role in activating the MAPK signaling pathway. We measured the phosphorylation of Erk (42kDa and 44 kDa) in C2C12 cells after treatment with cells with 9,10-dihydro-15d-PGJ2 (10 µM), a 15d-PGJ2 analog which is devoid of the electrophilic center, for 1 hr (after 24 hr. of serum starvation). We observed that the phosphorylation of Erk (42 kDa and 44 kDa) in C2C12 cells treated with 9,10-dihydro-15d-PGJ2 was significantly reduced (∼70%) as compared to the treatment with 15d-PGJ2 (Fig. 3F). This shows that 15d-PGJ2 activates the HRas-MAPK signaling pathway via the electrophilic center in its cyclopentenone ring.

15d-PGJ2 increases the localization of EGFP-tagged HRas at the plasma membrane compared to the Golgi in a C-terminal cysteine-dependent manner

15d-PGJ2 covalently modifies cysteine 184 and activates HRas signaling (Fig. 3). Reversible palmitoylation of cysteine 181 and cysteine 184 in the C-terminal tail of HRas regulate intracellular distribution and signaling of HRas. Inhibition of palmitoylation of the C-terminal cysteine 181, either by a palmitoylation inhibitor 2-Bromopalmitate or by mutation to serine, causes accumulation of HRas at the Golgi compared to the plasma membrane and alters activity(Rocks et al., 2005). Therefore, we tested whether the modification of 15d-PGJ2 alters the intracellular distribution of HRas. We co-expressed the EGFP-tagged wild type and the cysteine mutants of HRas (EGFP-HRas WT/C181S/C184S) with a previously reported marker of Golgi(Shaner et al., 2008) in C2C12 cells and stained the cells with plasma membrane marker WGA-633 (Fig. 4A). We compared Rmean, the ratio of mean EGFP-HRas intensity at the Golgi to the mean HRas intensity at the plasma membrane, to measure the distribution of HRas between the plasma membrane and the Golgi. We measured the intracellular distribution of HRas between the Golgi and the plasma membrane in C2C12 cells after treatment with 15d-PGJ2 (10 µM) for 24 hours in DMEM 10% serum medium and observed a significant decrease (∼20%) in the Rmean of C2C12 cells expressing the wild-type HRas after treatment with 15d-PGJ2 (Fig. 4B). However, we did not observe a change in the Rmean of C2C12 cells expressing HRas C181S or HRas C184S after treatment with 15d-PGJ2 (Fig. 4C). These observations suggest that 15d-PGJ2 increases the localization of HRas at the plasma membrane as compared to that in the Golgi in an HRas C-terminal cysteine-dependent manner.

15d-PGJ2 controls the intracellular distribution of HRas and differentiation of C2C12 cells in an HRas C-terminal cysteine-dependent manner.

A. Representative confocal micrograph of C2C12 myoblasts showing localization of EGFP-tagged HRas between the plasma membrane (stained with Alexa Fluor 633 conjugated Wheat Germ Agglutinin) and the Golgi (labelled with TagRFP-tagged Golgi resident GalT protein). A statistic Rmean was defined as the ratio of mean HRas intensity at the Golgi to the mean HRas intensity at the plasma membrane.

B. Distribution of the wild-type HRas between the Golgi and the plasma membrane, measured by Rmean, in C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO for 24 hours.

C. Distribution of the C-terminal cysteine mutants of HRas between the Golgi and the plasma membrane, measured by Rmean, in C2C12 myoblasts treated with 15d-PGJ2 (10 µM) or DMSO for 24 hours.

D. Expression of mRNAs of known markers of differentiation (MyoD, MyoG, and MHC), measured by qPCR, in differentiating C2C12 myoblasts expressing the EGFP-tagged wild-type and the C-terminal cysteine mutants of HRas after treatment with 15d-PGJ2 (4 µM) or DMSO for 5 days.

E. Expression of MHC protein, measured by immunoblotting, in differentiating C2C12 myoblasts expressing the EGFP-tagged wild-type and the C-terminal cysteine mutants of HRas after treatment with 15d-PGJ2 (4 µM) or DMSO for 5 days.

(Statistical significance was tested by the two-tailed student’s t-test ns=p>0.05, *=p<0.05, **=p<0.01, ***=p<0.001, ****=p<0.0001)

15d-PGJ2 mediated inhibition of differentiation of C2C12 cells is rescued by C181S and C184S mutants of HRas

HRas inhibits the differentiation of C2C12 myoblasts(Engler et al., 2021; Konieczny et al., 1989; Lassar et al., 1989; Lee et al., 2010; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007).15d-PGJ2 covalently modifies cysteine 184 and activates HRas (Fig. 3). Therefore, we tested whether the inhibition of myoblast differentiation by 15d-PGJ2 depends on the activation of HRas signaling by modification of the C-terminal cysteine 184. We expressed the wild-type and the cysteine mutants of HRas (EGFP-HRas WT/C181S/C184S) in C2C12 myoblasts and treated the cells with 15d-PGJ2 (4 µM) or DMSO during differentiation. We observed a decrease in the levels of mRNA of MHC in C2C12 cells expressing HRas WT and HRas C181S after 5 days of treatment with 15d-PGJ2. We did not observe this in expressing HRas C184S (Fig. 4D). We also observed a significant decrease in the protein levels of MHC in differentiating C2C12 cells expressing HRas WT and HRas C181S after treatment with 15d-PGJ2 (Fig. 4E). This decrease was partially rescued in cells expressing HRas C184S (Fig. 4E). These observations suggest that the inhibition of myoblast differentiation by 15d-PGJ2 depends on modification of HRas C-terminal cysteine 184.

Discussion

Senescence is characterized by an irreversible arrest in cell proliferation(Hayflick, 1965). Cells undergo senescence because of a myriad of stresses, including DNA damage, mitochondrial damage, and oncogene overexpression(Bihani et al., 2007, 2004; Casar et al., 2018; Chen and Ames, 1994; Chen et al., 1998; Coppé et al., 2008; D’Adda Di Fagagna, 2008; D’Adda Di Fagagna et al., 2003; Di Leonardo et al., 1994; Franza et al., 1986; Land et al., 1983; Robles and Adami, 1998; Serrano et al., 1997; Wiley et al., 2016; Woods et al., 1997). Senescent cells exhibit a multi-faceted physiological response, where they exhibit a flattened morphology, increase in cell size(Chen and Ames, 1994; Serrano et al., 1997), upregulation of tumor suppressor proteins(Calabrese et al., 2009; Lowe et al., 2004; Stein et al., 1990; Zindy et al., 2003), expression of neutral pH active β-galactosidase (SA β-gal)(Dimri et al., 1995; Lee et al., 2006), and altered metabolic state(Bittles and Harper, 1984; Jones et al., 2005; Wiley and Campisi, 2021, 2016; Zwerschke et al., 2003). Arachidonic acid metabolism is upregulated in senescent cells, which leads to increased synthesis of eicosanoid prostaglandins, which regulate the physiology of senescent cells(Wiley et al., 2021; Wiley and Campisi, 2021). Senescent cells exhibit a secretory phenotype (SASP) consisting of a variety of bioactive molecules including cytokines and chemokines, growth factors, matrix metalloproteases, etc(Coppé et al., 2008). Senescent cells influence the surrounding cells via the SASP factors, which regulate proliferation, migration, and other cell biological processes in the neighboring cells(Campisi, 2005). SASP-mediated perturbations in the microenvironment are implicated in several senescence-associate pathologies(Wiley and Campisi, 2021). Senescent fibroblasts increase the proliferation of premalignant and malignant epithelial cells(Krtolica et al., 2001). Conditioned medium of senescent fibroblasts promoted tumorigenesis in mouse keratinocytes(Dilley et al., 2003). Senescent fibroblasts transform pre-malignant breast cancer cells into invasive, tumor-forming cells(Parrinello et al., 2005). Senescence in muscle stem cells induces sarcopenia via activation of the p38 MAP kinase pathway and transient inhibition of the p38 MAP kinases rejuvenates aged muscle stem cells to ameliorate sarcopenia(Cosgrove et al., 2014). Senescent cells inhibit the differentiation of myoblasts by secretion of IL6 by senescent muscle stem cells in myotonic dystrophy(Conte et al., 2023).

In this study, we show that senescent myoblasts synthesize and release eicosanoid prostaglandin 15-deoxy-Δ12,14-prostaglandin J2 (15d-PGJ2) (Fig. 1I and J), the terminal non-enzymatic dehydration product of prostaglandin PGD2(Shibata et al., 2002). We used Doxorubicin (Doxo) to induce senescence in C2C12 myoblasts and showed that the conditioned medium of senescent C2C12 cells inhibits differentiation of C2C12 myoblasts (Fig. 2A). Inhibition of synthesis of PGD2 by treatment of senescent cells with AT-56, a well-characterized inhibitor of prostaglandin D synthase(Hu et al., 2021; S. Hu et al., 2023; Shunfeng Hu et al., 2023; Irikura et al., 2009), rescued this inhibitory effect of the conditioned medium on the differentiation of myoblasts (Fig. 2A). A study has shown that prostaglandin PGD2 inhibits differentiation of C2C12 myoblasts(Veliça et al., 2010), but the authors noted that knockout of DP1 and DP2 (the known receptors of prostaglandins PGD2) does not abrogate inhibition of differentiation of myoblasts by PGD2. This observation suggested that PGD2 might inhibit the differentiation of myoblasts by a receptor-independent mechanism, possibly by its spontaneous non-enzymatic dehydration to 15d-PGJ2. 15d-PGJ2 has been suggested to be an endogenous ligand of PPARγ(Li et al., 2019). However, the inhibition of PPARγ did not abrogate the inhibition of differentiation of C2C12 myoblasts by 15d-PGJ2, suggesting the existence of other possible mechanisms(Hunter et al., 2001). 15d-PGJ2 has varied effects on cell physiology in a context-dependent manner. On one hand, 15d-PGJ2 promotes tumorigenesis by inducing epithelial to mesenchymal transition in breast cancer cell line MCF7(Choi et al., 2020), 15d-PGJ2 inhibits the proliferation of A549, H1299, and H23 lung adenocarcinoma cells via induction of ROS and activation of apoptosis(Slanovc et al., 2024). Here, we show that 15d-PGJ2 inhibits the proliferation and the differentiation of C2C12 myoblasts (Fig. 2B, C and D).

15d-PGJ2 contains an electrophilic cyclopentenone ring in its structure, allowing 15d-PGJ2 to covalently modify and form Michael adducts with cysteine residues of proteins(Shibata et al., 2002). A previous proteomic study in endothelial cells showed biotinylated 15d-PGJ2 covalently modified over 300 proteins, which regulate several physiological processes including cell cycle (MAPK1, MCM4), cell metabolism (Fatty acid synthase, Isocitrate dehydrogenase), apoptosis (PDCD6I), translation (Elongation factor 1 and 2, EIF4A-I), intracellular transport (Importin subunit β1, Exportin 2, Kinesin 1 heavy chain)(Marcone and Fitzgerald, 2013). Another proteomic study in neuronal cells suggested that 15d-PGJ2 modifies several proteins including chaperone HSP8A, glycolytic proteins Enolase 1 and 2, GAPDH, PKM1, cytoskeleton proteins Tubulin β2b, β actin, GFAP, etc(Yamamoto et al., 2011). This study also showed modification of peptide fragments homologous to IκB kinase β, Thioredoxin, and a small molecule GTPase HRas. 15d-PGJ2 has been shown to covalently modify HRas in NIH3T3 and Cos1 cells(Luis Oliva et al., 2003) and IMR90 cells(Wiley et al., 2021). Modification by 15d-PGJ2 led to the activation of HRas, judged by an increase in GTP-bound HRas. It is clear that 15d-PGJ2 is capable of modifying numerous proteins in different contexts. Despite these observations, the functional relevance of these modifications in numerous contexts remains to be mapped. Here we focused on the role of 15d-PGJ2 in the context of senescence and skeletal muscle differentiation. In this study, we showed that 15d-PGJ2 covalently modifies HRas at cysteine 184 but not cysteine 181 in C2C12 myoblasts (Fig. 3A). We showed by FRET microscopy that modification of HRas by 15d-PGJ2 in HRas WT and HRas C181S activates HRas in C2C12 cells, but 15d-PGJ2 is unable to activate HRas C184S in this context (Fig. 3B, C, and D and Fig. S3A and B). This observation shows a direct link between the modification of HRas by 15d-PGJ2 and the activation of HRas GTPase.

HRas activates two major downstream signaling pathways, the HRas-MAPK and the HRas-PI3K pathway(Pylayeva-Gupta et al., 2011). We showed that covalent modification of HRas by 15d-PGJ2 via the electrophilic cyclopentenone ring activates HRas (Fig. 3C and D) and activates the HRas-MAPK pathway, demonstrated by an increase in the phosphorylation of Erk after treatment with 15d-PGJ2 (Fig. 3E and F). However, we did not observe activation of the HRas-PI3K pathway, as we did not see an increase in the phosphorylation of Akt after treatment with 15d-PGJ2 (Fig. S3C). MAPK and PI3K pathways are known regulators of muscle differentiation(Bennett and Tonks, 1997; Rommel et al., 1999), where inhibition of the RAF-MEK-Erk pathway or activation of the PI3K pathway promotes the differentiation of myoblasts. Preferential activation of the HRas-MAPK pathway over the HRas-PI3K pathway after treatment with 15d-PGJ2 can be a possible mechanism by which 15d-PGJ2 can inhibit the differentiation of myoblasts. HRas is known to regulate the differentiation of myoblasts in different contexts. Constitutively active HRas signaling by expression of oncogenic HRas mutant (HRas V12) leads to inhibition of differentiation of myoblasts(Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Van Der Burgt et al., 2007). Here we showed that the inhibition of differentiation of myoblasts after 15d-PGJ2 is partially rescued in cells expressing the C184S mutant of HRas but not the wild type or the C181S mutant (Fig. 4D and E and S4E). HRas C184S did not get modified by 15d-PGJ2 (Fig. 3A). These observations suggest that the inhibition of differentiation of myoblasts by 15d-PGJ2 is partially dependent on the covalent modification of HRas by 15d-PGJ2.

Cysteine 181 and 184 in the C-terminal of HRas regulate the intracellular distribution of HRas between the plasma membrane and the Golgi by reversible palmitoylation and de-palmitoylation(Rocks et al., 2005). Inhibition of the palmitoylation of C-terminal cysteine 181, either by treatment with protein palmitoylation inhibitor 2-bromopalmitate or mutation of cysteine to serine, leads to accumulation of HRas at the Golgi. Intracellular localization of HRas maintains two distinct pools of HRas activity, where the plasma membrane pool shows a faster activation followed by short kinetics and the Golgi pool shows a slower activation but a sustained activation(Agudo-Ibáñez et al., 2015; Busquets-Hernández and Triola, 2021; Lorentzen et al., 2010; Rocks et al., 2005). We showed that the covalent modification of HRas by 15d-PGJ2 alters the intracellular distribution of HRas. We showed that the covalent modification of HRas by 15d-PGJ2 leads to an increase in the localization of the wild-type HRas at the plasma membrane compared to the Golgi (Fig. 4B). We did not observe any changes in the intracellular distribution of HRas C181S or HRas C184S after treatment with 15d-PGJ2 (Fig. 4C). HRas C184S is not modified by 15d-PGJ2, but HRas C181S is modified by 15d-PGJ2 (Fig. 3A). This suggests that the intracellular redistribution of HRas due to covalent modification by 15d-PGJ2 at cysteine 184 requires palmitoylation of cysteine 181.

Previous reports suggest that downstream signaling of HRas depends on the intracellular localization of HRas(Rocks et al., 2005; Santra et al., 2019). For example, targeted localization of HRas at the ER membrane induced expression of cell-migration genes. Localization of HRas at the plasma membrane showed a strong correlation with the expression of cell cycle genes, particularly the MAPK signaling pathway. Localization of HRas at the plasma membrane also showed a negative correlation with genes associated with the PI3K-Akt pathway. Here we showed that the intracellular distribution of HRas regulates differentiation of myoblasts. In order to show this, we used the constitutively active mutant of HRas (HRas V12) which has been shown to inhibit the differentiation of myoblasts(Engler et al., 2021; Konieczny et al., 1989; Lassar et al., 1989; Olson,’ et al., 1987; Scholz et al., 2009; Van Der Burgt et al., 2007). We expressed cysteine mutants of HRas V12 in C2C12 myoblasts and found that HRas V12 C181S localized predominantly at the Golgi whereas HRas V12 and HRas V12 C184S localized at both the plasma membrane and the Golgi (Fig. S4A). When differentiated, we observed that C2C12 cells expressing HRas V12 C181S differentiated but HRas V12 or HRas V12 C184S did not differentiate (Fig. S4B, C, and D). These observations suggest alteration of intracellular distribution of HRas affects the HRas-mediated inhibition of the differentiation of myoblasts.

Doxorubicin (Doxo) is a widely used chemotherapy agent for the treatment of cancers(Johnson-Arbor and Dubey, 2022). Treatment with Doxo induces senescence. Doxo-mediated DNA damage leads to p53, p16, and p21-dependent senescence in human fibroblasts(Di Leonardo et al., 1994; Robles and Adami, 1998). On the other hand, treatment with doxorubicin leads to a decrease in muscle mass and cross-sectional area, leading to chemotherapy-induced cachexia(Hiensch et al., 2020). Several mechanisms have been proposed behind chemotherapy-induced cachexia, including the generation of reactive oxygen species(Gilliam and St. Clair, 2011), activation of proteases like calpain and caspases(Gilliam et al., 2012; Smuder et al., 2011), and impaired insulin signaling(de Lima Junior et al., 2016). This study provides a possible mechanism behind chemotherapy-induced loss of muscle mass and functioning. Induction of senescence in myoblasts by treatment with Doxo could lead to increased synthesis and release of 15d-PGJ2 by senescent cells which could be taken up by myoblasts in the microenvironment. The lipid could covalently modify and activate HRas at cysteine 184 to inhibit the differentiation of myoblasts. Therefore, targeting the synthesis and release of 15d-PGJ2 by senescent cells could serve as an important target to promote skeletal muscle homeostasis in cancer patients.